# Variables

#### Learning goals {-}

* How to define variables and work with them in `R`  
* Learn the various possible classes of data in `R`  

## Introducing variables {-}

So far we have strictly been using `R` as a calculator, with commands such as: 

```{r,echo=TRUE,collapse=TRUE}
3 + 5
```

Of course, `R` can do much, much more than these basic computations. Your first step in uncovering the potential of `R` is learning how to use **variables**.

In `R`, a variable is a convenient way of referring to an underlying value. That value can be as simple as a single number (e.g., `6`), or as complex as a spreadsheet that is many Gigabytes in size. It may be useful to think of a variable as a cup; just as cups make it easy to hold your coffee and carry it from the kitchen to the couch, variables make it easy to contain and work with data.

### Declaring variables {-}

To assign numbers or other types of data to a variable, you use the `<` and `-` characters to make the arrow symbol `<-`.

```{r,echo=TRUE,collapse=TRUE}
x <- 3+5
```

As the direction of the `<-` arrow suggests, this command stores the result of `3 + 5` into the variable `x`.

Unlike before, you did not see `8` printed to the *Console*. That is because the result was stored into `x`.


### Calling variables {-}

If you wanted R to tell you what `x` is, just type the variable name into the *Console* and run that command:

```{r,echo=TRUE,collapse=TRUE}
x
```

Want to create a variable but also see its value at the same time? Here's a handy trick: put your command in parentheses:

```{r,echo=TRUE,collapse=TRUE}
(x <- 3*12)
```

When you do that, `x` gets assigned a value, then that value is printed to the console.

You can also update variables.

```{r,echo=TRUE,collapse=TRUE}
(x <- x * 3)
```

```{r,echo=TRUE,collapse=TRUE}
(x <- x * 3)
```

You can also add variables together.

```{r,echo=TRUE,collapse=TRUE}
x <- 8
y <- 4.5
x + y
```

### Naming variables  {-}

Here are a few rules:  

**1.** A variable name has to have at least one letter in it. These examples work:

**2.** A variable name has to be connected. No spaces! It is usually best to represent a space using a period (`.`) or an underscore (`_`). Note that periods and underscores can be used in variable names:

```{r,echo=TRUE,collapse=TRUE}
my.variable <- 5 # periods can be used
my_variable <- 5 # underscores can be used
```

However, hyphens *cannot* be used, since that symbol is used for subtraction.

**3.** Variables are case-sensitive. If you misspell a variable name, you will confuse `R` and get an error.  For example, ask `R` to tell you the value of capital `X`. The error message will be `Error: object 'X' not found`, which means `R` looked in its memory for an object (i.e., a variable) named `X` and could not find one.

**4.** Variable names can be as complicated or as simple as you want. 

**5.** Some names need to be avoided, since `R` uses them for special purposes. For example, `data` should be avoided, as should `mean`, since both are functions built-in to `R` and `R` is liable to interpret them as such instead of as a variable containing your data.

```{r,echo=TRUE,collapse=TRUE}
supercalifragilistic.expialidocious <- 5
supercalifragilistic.expialidocious  # still works
```

So those are the basic rules, but naming variables well is a bit of an art. The trick is using names that are clear but are not so complicated that typing them is tedious or prone to errors.

Note that `R` uses a feature called 'Tab complete' to help you type variable names. Begin typing a variable name, such as `supercalifragilistic.expialidocious` from the example above, but after the first few letters press the `Tab` key. `R` will then give you options for auto-completing your word. Press `Tab` again, or `Enter`, to accept the auto-complete. This is a handy way to avoid typos.



## Types of data in `R` {-}

So far we have been working exclusively with numeric data. But  there are many different data types in R. We call these “types” of data **classes**:

* Decimal values like 4.5 are called **numeric** data.
* Natural numbers like 4 are called **integers**. Integers are also numerics.
* Boolean values (TRUE or FALSE) are called **logical** data.
* Text (or string) values are called **character** data.

In order to be combined, data have to be the same class. 

R is able to compute the following commands ...

```{r,echo=TRUE,collapse=TRUE}
x <- 6
y <- 4
x + y
```

... but not these:

```{r eval=FALSE,echo=TRUE,collapse=TRUE}
x <- 6
y <- "4"
x + y
```

That's because the quotation marks used in naming `y` causes `R` to interpret `y` as a `character` class.

To see how `R` is interpreting variables, you can use the `class()` function:

```{r,echo=TRUE,collapse=TRUE}
x <- 100
class(x)
```

```{r,echo=TRUE,collapse=TRUE}
x <- "100"
class(x)
```

```{r,echo=TRUE,collapse=TRUE}
x <- 100 == 101
class(x)
```

Another data type to be aware of is **factors**, but we will deal with them later.

#### Exercises {-}

**Finding the errors**  

**1.** This code will produce an error. Can you find the problem and fix it so that this code will work?

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
# Assign 5 to a variable
my_var < 5
```

**2.** Same for this one:

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
# Assign 5 to a variable
my_var == 5
```

**3.** Same for this one:
```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
x <- 5
y <- 1
X + y
```

&nbsp;  

**Your Bananas-to-ICS ratio**

**4.** Estimate how many bananas you've eaten in your lifetime and store that value in a variable (choose whatever name you wish). (By the way, what is a good method for estimating this as accurately as you can?)

**5.** Now estimate how many ice cream sandwiches you've eaten in your lifetime and store that in a different variable.  

**6.** Now use these variables to calculate your Banana-to-ICS ratio. Store your result in a third variable, then call that variable in the Console to see your ratio.  

**7.** Who in the class has the highest ratio? Who has the lowest?  

&nbsp;  


**Creating boolean variables**

**8.** Assign a `FALSE` statement of your choosing to a variable of whatever name you wish.  

**9.** Confirm that the class of this variable is “logical.”  

**10.** Confirm that the variable equals `FALSE`.  

&nbsp;  


**Converting Fahrenheit to Celsius:**  

**11.** Assign a variable `fahrenheit` the numerical value of 32.  

**12.** Assign a variable `celsius` to equal the conversion from Fahrenheit to Celsius. Unless you're a meteorology nerd, you may need to Google the equation for this conversion. 

**13.** Print the value of `celsius` to the *Console*.  

**14.** Now use this code to determine the *Celsius* equivalent of 212 degrees *Fahrenheit*.

&nbsp;  

**Wrapping up**

**15.** Now ensure that your entire script is properly commented, and make sure your script is saved in your `datalab` working directory before closing.
