# Structures for data in `R`

```{r,echo=FALSE,eval=TRUE}
# Use this for developing content; make sure eval==FALSE before you deploy.
#is_teacher=FALSE
```

#### Learning goals {-}

* Learn the various structures of data in `R`  
* How to work with vectors in `R`  


```{r, results = 'asis', eval = is_teacher,echo=FALSE}
teacher_content('Here is some teacher content.')
```


## Introducing data structures {-}

Data belong to different *classes*, as explained in the previous module, and they can be arranged into various **structures**. 

So far we have been dealing only with variables that contain a single value, but the real value of `R` comes from assigning *entire sets* of data to a variable. 

## Vectors {-}

The simplest data structure in `R` is a **vector**. A vector is simply a set of values. A vector can contain only a single value, as we have been working with thus far, or it can contain many millions of values. 

### Declaring and using vectors {-}

To build up a vector in `R`, use the function `c()`, which is short for "concatenate".

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- c(5,6,7,8)
x
```

You can use the `c()` function to concatenate two vectors together:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- c(5,6,7,8)
y <- c(9,10,11,12)
z <- c(x,y)
z
```

You can also use `c()` to add values to a vector:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- c(5,6,7,8)
x <- c(x,9)
x
```

When two vectors are of the same length, you can do arithmetic with them:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- c(5,6,7,8)
y <- c(9,10,11,12)
x + y
```

```{r,echo=TRUE,collapse=TRUE}
x - y
```

```{r,echo=TRUE,collapse=TRUE}
x * y
```

```{r,echo=TRUE,collapse=TRUE}
x / y
```


You can also put vectors through logical tests:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- 1:5
4 == x
```

This command is asking `R` to tell you whether each element in `x` is equal to `4`. 

You can create vectors of any data class (i.e., data type).

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- c("Ben","Joe","Eric") 
x
```

```{r,echo=TRUE,collapse=TRUE}
y <- c(TRUE,TRUE,FALSE)
y
```

Note that all values within a vector *must* be of the same class. You can't combine numerics and characters into the same vector. If you did, `R` would try to convert the numbers to characters.  For example:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- 4
y <- "6"
z <- c(x,y)
z
```

#### Exercise 1 {-}

A. Create a vector with at least one number, character (string) and logical value 

B. Identify the class of each element of the vector

### Useful functions for handling vectors  {-}

**`length()`** tells you the number of elements in a vector:

```{r,echo=TRUE,collapse=TRUE}
x <- c(5,6)
length(x)
```

```{r,echo=TRUE,collapse=TRUE}
y <- c(9,10,11,12)
length(y)
```

The **colon symbol `:`** creates a vector with every integer occurring between a min and max:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- 1:10
x
```

**`seq()`** allows you to build a vector using evenly spaced *sequence* of values between a min and max:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
seq(0,100,length=11)
```

In this command, you are telling `R` to give you a sequence of values from `0` to `100`, and you want the length of that vector to be `11`. `R` then figures out the spacing required between each value in order to make that happen. 

Alternatively, you can prescribe the interval between values instead of the length:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
seq(0,100,by=7)
```

**`rep()`** allows you to repeat a single value a specified number of times:

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
rep("Hey!",times=5)
```

**`head()`** and **`tail()`** can be used to retrieve the first 6 or last 6 elements in a vector, respectively.

```{r,echo=TRUE,collapse=TRUE,collapse=TRUE}
x <- 1:1000
head(x)
tail(x)
```

You can also adjust how many elements to return:

```{r,echo=TRUE,collapse=TRUE}
head(x,2)
tail(x,10)
```

**`sort()`** allows you to order a vector from its smallest value to its largest:

```{r,echo=TRUE,collapse=TRUE}
x <- c(4,8,1,6,9,2,7,5,3)
sort(x)
```

**`rev()`** lets you reverse the order of elements within a vector:

```{r,echo=TRUE,collapse=TRUE}
x <- c(4,8,1,6,9,2,7,5,3)
rev(x)
```

```{r,echo=TRUE,collapse=TRUE}
rev(sort(x))
```

**`which()`** allows you to ask, "For which elements of a vector is the following statement true?"

```{r,echo=TRUE,collapse=TRUE}
x <- 1:10
which(x==4)
```

If no values within the vector meet the condition, a vector of length zero will be returned:

```{r,echo=TRUE,collapse=TRUE}
x <- 1:10
which(x == 11)
```

**`%in%`** is a handy operator that allows you to ask whether a value occurs *within* a vector:

```{r,echo=TRUE,collapse=TRUE}
x <- 1:10
4 %in% x
```

```{r,echo=TRUE,collapse=TRUE}
11 %in% x
```

#### Exercise 2 {-}

A. Use the colon symbol to create a vector of length 5 between a minimum and a maximum value of your choosing.  

B. Create a second vector of length 5 using the `seq()` function.  

C. Create a third vector of length 5 using the `rep()` function. 

D. Finally, concatenate the three vectors and check that the length makes sense.  


### Subsetting vectors {-}

Since you will eventually be working with vectors that contain thousands of data points, it will be useful to have some tools for *subsetting* them -- that is, looking at only a few select elements at a time.

You can subset a vector using square brackets `[ ]`.

```{r,echo=TRUE,collapse=TRUE}
x <- 50:100
x[10]
```

This command is asking `R` to return the 10th element in the vector `x`.

```{r,echo=TRUE,collapse=TRUE}
x[10:20]
```

This command is asking `R` to return elements `10:20` in the vector `x`. 

#### Exercise 2 {-}

A. Figure out how to replicate the `head()` function using your new vector subsetting skills.

```{r,eval=is_teacher,echo=is_teacher,collapse=TRUE}
x[1:6]
```

B. Now replicate the `tail()` function, using those same skills as well as the `length()` function you just learned.

```{r,eval=is_teacher,echo=is_teacher,collapse=TRUE}
x[(length(x) - 5) : length(x)]
```


#### Review assignment {-}

A. Create a vector called "sleep_time" with the number of hours you slept for each day in the last week.

B. Check if you slept more on day 3 than day 7

C. Get the total number of hours slept in the last week

D. Get the average number of hours slept in the last week

E. Check if the total number of hours in the first 3 days is less than the total number of hours in the last 4 days

F. Create an object named `over_under`. This should be the difference between how much you slept each night and 8 hours (ie, 1.5 means you slept 9.5 hours and -2 means you slept 8 hours)

G. Write code to use `over_under` to calculate your sleep deficit / surplus this week (ie, the total hours over/under the amount of sleep you would have gotten had you slept 8 hours every night)

H. Write code to get the minimum number of hours you slept this week.

I. Write code to calculate how many hours of sleep you would have gotten had you sleep the minimum number of hours every night.

J. Write code to calculate the average of the hours of sleep you got on the 3rd through 6th days of the week.

K. Write code to calculate how many hours of sleep you would get in a year if you were to sleep the same amount every night as the average amount you slept from the 3rd to 6th days of the week.

L. Write code to calculate how many hours of sleep per year someone who sleeps 8 hours a night gets.

M. How many hours more/less than the 8 hours per night sleeper do you get in a year, assuming you sleep every night the average of the amount you slept on the first and last day of this week?

