# Structures for data in `R`

```{r,echo=FALSE,eval=FALSE}
# Use this for developing content; make sure eval==FALSE before you deploy.
is_teacher=FALSE
```

## Learning goals {-}

* Learn the various structures of data in `R`  
* How to work with vectors in `R`.  


```{r, results = 'asis', eval = is_teacher,echo=FALSE}
teacher_content('Here is some teacher content.')
```


## Introducing data structures {-}

Data belong to different *classes*, as explained in the previous module, and they can be arranged into various **structures**. 

So far we have been dealing only with variables that contain a single value, but the real value of `R` comes from assigning *entire sets* of data to a variable. 

## Vectors {-}

The simplest data structure in `R` is a **vector**. A vector is simply a set of values. A vector can contain only a single value, as we have been working with thus far, or it can contain many millions of values. 

### Declaring and using vectors {-}

To build up a vector in `R`, use the function `c()`, which is short for "concatenate".

```{r}
x <- c(5,6,7,8)
x
```

You can use the `c()` function to concatenate two vectors together:

```{r}
x <- c(5,6,7,8)
y <- c(9,10,11,12)
z <- c(x,y)
z
```

You can also use `c()` to add values to a vector:

```{r}
x <- c(5,6,7,8)
x <- c(x,9)
x
```

When two vectors are of the same length, you can do arithmetic with them:

```{r}
x <- c(5,6,7,8)
y <- c(9,10,11,12)

x + y
x - y
x * y
x / y
```

You can also put vectors through logical tests:

```{r}
x <- 1:5
4 == x
```

This command is asking `R` to tell you whether each element in `x` is equal to `4`. 

You can create vectors of any data class:

```{r}
x <- c("Ben","Joe","Eric") 
x

y <- c(TRUE,TRUE,FALSE)
y
```

Note that all values within a vector *must* be of the same class (i.e., data type). You can't combine numerics and characters into the same vector. If you did, `R` would try to convert the numbers to characters.  For example:

```{r}
x <- 4
y <- "6"
z <- c(x,y)
z
```


### Useful functions for handling vectors  {-}

**`length()`** tells you the number of elements in a vector:

```{r}
x <- c(5,6)
y <- c(9,10,11,12)

length(x)
length(y)
```

The **colon symbol `:`** creates a vector with every integer occurring between a min and max:

```{r}
x <- 1:10
x
```

**`seq()`** allows you to build a vector using evenly spaced *sequence* of values between a min and max:

```{r}
seq(0,100,length=11)
```

In this command, you are telling `R` to give you a sequence of values from `0` to `100`, and you want the length of that vector to be `11`. `R` then figures out the spacing required between each value in order to make that happen. 

Alternatively, you can prescribe the interval between values instead of the length:

```{r}
seq(0,100,by=7)
```

**`head()`** and **`tail()`** can be used to retrieve the first 6 or last 6 elements in a vector, respectively.

```{r}
x <- 1:1000
head(x)
tail(x)
```

You can also adjust how many elements to return:

```{r}
head(x,2)
tail(x,10)
```

**`which()`** allows you to ask, "For which elements of a vector is the following statement true?"

```{r}
x <- 1:10
which(x==4)
```

If no values within the vector meet the condition, a vector of length zero will be returned:

```{r}
x <- 1:10
which(x == 11)
```

**`%in%`** is a handy operator that allows you to ask whether a value occurs *within* a vector:

```{r}
x <- 1:10
4 %in% x
11 %in% x
```

#### Exercise 2 {-}

*NOTE: UNDER CONSTRUCTION!*


### Subsetting vectors {-}

Since you will eventually be working with vectors that contain thousands of data points, it will be useful to have some tools for *subsetting* them -- that is, looking at only a few select elements at a time.

You can subset a vector using square brackets `[ ]`.

```{r}
x <- 50:100
x[10]
```

This command is asking `R` to return the 10th element in the vector `x`.

```{r}
x[10:20]
```

This command is asking `R` to return elements `10:20` in the vector `x`. 

#### Exercise 3 {-}

A. Figure out how to replicate the `head` function using your new vector subsetting skills.

```{r,eval=is_teacher}
x[1:6]
```

B. Now replicate the `tail()` function, using those same skills as well as the `length()` function you just learned.

```{r,eval=is_teacher}
x[(length(x) - 5) : length(x)]
```

### Dataframes & other data structures  {-}

A **vector** is the most basic data structure in `R`, and the other structures are built out of vectors.

As a data scientist, the most common data structure you will be working with is a **dataframe**, which is essentially a spreadsheet: a dataset with rows and columns, in which each column represents is a vector of the same class of data. 

We will explore dataframes in detail later, but here is a sneak peak at what they look like:

```{r}
df <- data.frame(x=300:310,
           y=600:610)
df
```

In this command, we used the `data.frame()` function to combine two vectors into a dataframe with two columns named `x` and `y`. `R` then saved this result in a new variable named `df`. When we call `df`, `R` shows us the dataframe.

The great thing about dataframes is that they allow you to relate different data types to each other.

```{r}
df <- data.frame(name=c("Ben","Joe","Eric"),
                 height.inches=c(75,73,80))
df
```

This dataframe has one column of class `character` and another of class `numeric`.

The two other most common data structures are **matrices** and **lists**, but we will wait on learning about thos. For now, focus on becoming comfortable using vectors and dataframes. 


### Exercise 3 {-}

*NOTE: UNDER CONSTRUCTION!*


## Review assignment {-}

*NOTE: UNDER CONSTRUCTION!*


## Other Resources {-}




