# Basic subsetting and filtering {#subsetting}

```{r,echo=FALSE,eval=FALSE}
is_teacher <- FALSE
```

#### Learning goals {-}

* Understand how to subset / filter data

## Subsetting with indices {-}

You have already learned that certain elements of a vector can be called by specifiying an index:

```{r, echo = T,collapse=T}
x <- 55:65

# Call x without subsetting
x
```

```{r, echo = T,collapse=T}
# Now call only the third element of x
x[3]
```

You can also subset an object by calling multiple indices:

```{r, echo = T,collapse=T}
# Now call the third, fourth, and fifth element of x
x[c(3,4,5)]
```

```{r, echo = T,collapse=T}
# Another way of doing the same thing:
x[3:5]
```

## Subsetting with booleans {-}

You can also subset objects with 'booleans'. Recall that boolean / logical data have two possible values: `TRUE` or `FALSE`. For example:

```{r, echo = T,collapse=T}
joes_age <- 35
old_age <- 36
joe_is_old <- joes_age >= old_age
joe_is_old
```

Recall also that you can calculate whether a condition is `TRUE` or `FALSE` on multiple elements of a vector. For example:

```{r, echo = T,collapse=T}
ages <- c(10, 20, 30, 40, 50, 60)
old_age <- 36
people_are_old <- ages >= old_age
people_are_old
```

Boolean vectors are useful for subsetting. Think of 'subsetting' as keeping only those elements of a vector for which a condition is `TRUE`. 

```{r, echo = T,collapse=T}
x <- 55:59

# Call x without subsetting
x
```

```{r, echo = T,collapse=T}
# Now subset to the second, third, and fourth element
x[c(FALSE, TRUE, TRUE, TRUE, FALSE)]
```

That command returned elements for which the subetting vector was `TRUE`. 

This is equivalent to...

```{r, echo = T,collapse=T}
x[2:4]
```

You can also get the same result using a logical test, since logical tests return boolean values:

```{r, echo = T,collapse=T}
# Develop your logical test
x %in% c(56,57,58)

# Plug it into the subsetting brackets
x[ x %in% c(56,57,58) ]
```

This methods gets really useful when you are working with bigger datasets, such as this one:

```{r, echo = T,collapse=T}
# Make a large dataset of random numbers
y <- runif(n=100,
           min=0,
           max=10)
length(y)
```

```{r, echo = T,collapse=T}
range(y)
```

With a dataset like this, you can use a boolean filter to figure out how many values are greater than, say, 9.

First, develop your logical test, which will tell you whether each value in the vector is greater than 9:

```{r, echo = T,collapse=T}
# Develop your logical test,
y > 9
```

Now, to get the values corresponding to each `TRUE` in this list, plug your logical test into your subsetting brackets.

```{r, echo = T,collapse=T}
y[y > 9]
```

Here's another way you can do the same thing:

```{r, echo = T,collapse=T}
verdicts <- y > 9
y[verdicts]
```

You can use double logical tests too. For example, what if you want all elements between the values 7.0 and 9.0?

```{r, echo = T,collapse=T}
verdicts <- y > 7 & y < 9
y[verdicts]
```

### Review assignment {-}

A. Create a vector named `nummies` of all numbers from 1 to 100

B. Create another vector named `little_nummies` which consists of all those numbers which are less than or equal to 30

C. Create a boolean vector named `these_are_big` which indicates whether each element of `nummies` is greater than or equal to 70

D. Use `these_are_big` to subset `nummies` into a vector named `big_nummies`

E. Create a new vector named `these_are_not_that_big` which indicates whether each element of `nummies` is greater than 30 and less than 70. You'll need to use the `&` symbol.

F. Create a new vector named `meh_nummies` which consists of all `nummies` which are greater than 30 and less than 70.

G. How many numbers are greater than 30 and less than 70?

H. What is the sum of all those numbers in `meh_nummies`
