# Subsetting and filtering {#subsetting}

```{r,echo=FALSE,eval=FALSE}
is_teacher <- FALSE
```

#### Learning goals {-}

* Understand how to subset / filter


```{r, results = 'asis', eval = is_teacher}
teacher_content('Here is some teacher content.')
```

## A quick review of booleans {-}

Recall that a boolean / logical data type is either `TRUE` or `FALSE`. For example:

```{r}
joes_age <- 35
old_age <- 36
joe_is_old <- joes_age >= old_age
joe_is_old
```

Recall also that you can calculate whether a condition is true or false on multiple elements of a vector. For example:

```{r}
ages <- c(10, 20, 30, 40, 50, 60)
old_age <- 36
people_are_old <- ages >= old_age
people_are_old
```

Boolean vectors are useful for subsetting, ie keeping those elements of a vector for which a condition is `TRUE`. You have already done this using indexes, for example:

```{r}
ages[3]
```

But you can also do this using booleans. For example:

```{r}
foo <- 1:4
foo[2:4]
foo[c(2,3,4)]
foo[c(FALSE, TRUE, TRUE, TRUE)]
foo[foo >= 2]
```


### Review assignment {-}

A. Create a vector named `nummies` of all numbers from 1 to 100

B. Create another vector named `little_nummies` which consists of all those numbers which are less than or equal to 30

C. Create a boolean vector named `these_are_big` which indicates whether each element of `nummies` is greater than or equal to 70

D. Use `these_are_big` to subset `nummies` into a vector named `big_nummies`

E. Create a new vector named `these_are_not_that_big` which indicates whether each element of `nummies` is greater than 30 and less than 70. You'll need to use the `&` symbol.

F. Create a new vector named `meh_nummies` which consists of all `nummies` which are greater than 30 and less than 70.

G. How many numbers are greater than 30 and less than 70?

H. What is the sum of all those numbers in `meh_nummies`
