# Basics of `ggplot`

(Refer heavily to https://ggplot2-book.org/introduction.html)

#### Learning goals {-}

- Understand what `ggplot2` is and why it's used
- Be able to think conceptually in the framework of the "grammar of graphics"
- Learn the syntax for creating different plots using using `ggplot2`

## Thinking about data visualization

Let's start with a video: https://youtu.be/5Zg-C8AAIGg.

And one by Hans Rosling too: https://www.youtube.com/watch?v=FACK2knC08E

## What is `ggplot`?  {-}

`ggplot2` is an R package. It's one of the most downloaded packages in the R universe, and has become the gold standard for data visualization. It's extremely powerful and flexible, and allows for creating lots of visualizations of different types, ranging from maps to bare-bones academic publications, to complex, paneled charts with labeling, etc. Because the syntax is so different from "base" R, it can give the impression of having a somewhat steep learning curve. But in reality, because the principles are so conceptually simple, learning is fairly fast. Generally those who choose to learn it stick with it; that is, once you go gg, you don't go back.

## The name and concept  {-}

"GG" stands for "grammar of graphics", with "grammar" meaning "the fundamental principles or rules of an art or science" [@layered-grammar]. The most well-known "grammar of graphics" was written in 2005 and laid out some abstract principles for describing statistical graphics [@10.5555/1088896]. The basic idea is that all graphs can be described using a _layered_ grammar, and that all graphs have the same general elements...

- data
- aesthetics (mapping) of variables to objects
- geometric objects

... whereas some graphs have additional elements...

- statistical transformations
- labs
- facets
- themes


## Practice {-}

Titanice data set 

```{r}
library(ggplot2)
library(readr)
library(ggthemes)
library(dplyr)
titanic<- read_csv("https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/deaths.csv")

```

## Scatter plot {-} 

Call the plot

```{r}
ggplot()
```

Aesthetics (mapping variables to the x y plane)

```{r}
ggplot(data = titanic, aes(x = Age, y = Fare))
```

Geometry 

```{r}
ggplot(data = titanic, aes(x = Age, y = Fare)) + geom_point()
```

Statistics 
```{r}
ggplot(data = titanic, aes(x = Age, y = Fare)) + geom_point() + geom_smooth(method = 'lm')
```

Labs
```{r}
ggplot(data = titanic, aes(x = Age, y = Fare)) + geom_point() + geom_smooth(method = 'lm') + labs(title = 'Relationship between Age and Fare')
```

Facets 

```{r}
ggplot(data = titanic, aes(x = Age, y = Fare)) + geom_point() + geom_smooth(method = 'lm') + labs(title = 'Relationship between Age and Fare') +
  facet_wrap(~Sex)

```

Themes 

```{r}
ggplot(data = titanic, aes(x = Age, y = Fare)) + geom_point() + geom_smooth(method = 'lm') + labs(title = 'Relationship between Age and Fare') +
  facet_wrap(~Sex) + theme_fivethirtyeight()

```

## Barplot {-}

Get counts for sex
```{r}
ggplot(data = titanic, aes(x = Sex)) + geom_bar(stat='count')
```

Explicitly set y to another variable
```{r}
titanic_age <- titanic %>% group_by(Sex) %>% summarise(mean_age = mean(Age, na.rm= TRUE))
ggplot(data = titanic_age, aes(x = Sex,y= mean_age)) + geom_bar(stat = 'identity')
```

Fill
```{r}
ggplot(data = titanic_age, aes(x = Sex,y= mean_age)) + geom_bar(stat = 'identity', fill = 'blue', alpha = 0.5) 
```

Labs
```{r}

ggplot(data = titanic_age, aes(x = Sex,y= mean_age)) + geom_bar(stat = 'identity', fill = 'blue', alpha = 0.5) + labs(y = 'Mean age', title = 'Average age by sex')

```

Add another variable
```{r}
titanic_em <- titanic %>% group_by(Sex, Survived) %>% summarise(mean_age= mean(Age, na.rm = TRUE))
titanic_em <- titanic_em %>% mutate(Survived = ifelse(Survived == 1, 'Survived','Dead' ))
ggplot(data = titanic_em, aes(x=Sex, y=mean_age, fill = Survived)) + geom_bar(stat='identity')

```

Dodge
```{r}
ggplot(data = titanic_em, aes(x=Sex, y=mean_age, fill = Survived)) + geom_bar(stat='identity', position = 'dodge') 
```

Scales
```{r}
ggplot(data = titanic_em, aes(x=Sex, y=mean_age, fill = Survived)) + geom_bar(stat='identity', position = 'dodge', alpha = 0.7) + scale_fill_manual(values = c('blue', 'grey'))
```


## Another example 

```{r}
library(babynames)
bn <- babynames

# subset by the name Francis
francis <- bn %>% filter(name == 'Francis', sex == 'M')

# line plot
ggplot(data=francis, aes(x = year, y =n )) + geom_line() + labs(x = 'Year', y = 'Number of boys named Francis')

# layer with points
ggplot(data=francis, aes(x = year, y =n )) + geom_line() + geom_point() + labs(x = 'Year', y = 'Number of boys named Francis')

```

## A practical example {-}

Let's get practical (we'll get back to the theory later). 

First, let's read in some data on health from the World Bank:

```{r}
library(readr)
health <- read_csv('https://github.com/ZeningQu/World-Bank-Data-by-Indicators/raw/master/health/health.csv')
```

Canvas
Canvas + variables (mapping)
Canvas + variables (mapping) + geometric objects

## Learning examples {-}

### Perfecting the canvas {-}

Adjust y / x limits

Add a different background

Change x / y labels

### Aesthetic attributes of the geoms {-}


A scatterplot


Add a line of best fit



Add title / subtitle / caption

#### Review assignment  {-}

_Note: Under construction!_

## Other resources  {-}

_Note: Under construction!_


