# Dataframes: Intro 

#### Learning goal {-}

* Practice exploring, summarizing, and filtering dataframes

&nbsp;  

A vector is the most basic data structure in `R`, and the other structures are built out of vectors. But, as a data scientist, the most common data structure you will be working with -- by far -- is a **dataframe**. A dataframe, essentially, is a spreadsheet: a dataset with rows and columns, in which each column represents is a vector of the same class of data. 

## Introducing dataframes {-} 

Here is what a dataframe looks like:

```{r,echo=TRUE,collapse=TRUE}
# Using one of R's built-in datasets
head(iris)
```

In this dataframe, each row pertains to a unique iris plant. The columns contain related information about each individual plant. 

Here's another data.frame, built from scratch, which shows that dataframes are just a group of vectors:

```{r,echo=TRUE,collapse=TRUE}
x <- 25:29
y <- 55:59
df <- data.frame(x,y)
df
```

In this command, we used the `data.frame()` function to combine two vectors into a dataframe with two columns named `x` and `y`. `R` then saved this result in a new variable named `df`. When we call `df`, `R` shows us the dataframe.

The great thing about dataframes is that they allow you to relate different data types to each other.

```{r,echo=TRUE,collapse=TRUE}
df <- data.frame(name=c("Ben","Joe","Eric"),
                 height=c(75,73,80))
df
```

This dataframe has one column of class `character` and another of class `numeric`.


### Exercise 1 {-}

Let's create a new object named `animals`. This is going to be a dataframe with 4 different columns: `species`, `height`, `color`, `veg` (whether or not the animal is a vegetarian). 

First, come up with five species to add to your dataframe and list them in a vector named `species`. 

Second, make the other vectors with details about those species in the correct order. 

Third, combine these vectors into a dataframe named `animals`.


## Subsetting & exploring dataframes {-} 

To explore dataframes, let's use a dataset on fuel mileage for all cars sold from 1985 to 2014. 

```{r,echo=TRUE,collapse=TRUE}
# need to install first install.packages('fueleconomy')
library(fueleconomy)
data(vehicles)
head(vehicles)
```

To look at this dataframe in full, you call display it in a separate tab within `RStudio` using the `View()` functions:

```{r,echo=TRUE,collapse=TRUE,eval=FALSE}
View(vehicles)
```

A dataframe has rows of data organized into columns. In this dataframe, each row pertains to a single vehicle make/model -- i.e., a single *observation*. Each column pertains to a single *type* of data. Columns are named in the *header* of the dataframe. 

All the same useful exploration and [subsetting functions](#subsetting) that applied to vectors now apply to dataframes. In addition to those functions you already know, let's add some new functions to your inventory of useful functions.

### Exploration {-}

```{r,echo=TRUE,collapse=TRUE}
# head() and tail() summarize the beginning and end of the object
tail(vehicles) 
```

```{r,echo=TRUE,collapse=TRUE}
# get names of columns
names(vehicles) 
```

```{r,echo=TRUE,collapse=TRUE}
# get number of rows
nrow(vehicles) 
```

```{r,echo=TRUE,collapse=TRUE}
# get number of columns
ncol(vehicles) 
```

Note that `length()` does not work the same on dataframes as it does with vectors. In dataframes, `length()` is the equivalent of `ncol()`; it will *not* give you the number of rows in a dataset.

### Checking for NAs {-}

```{r,echo=TRUE,collapse=TRUE, eval=FALSE}
# is.na(vehicles)
any(is.na(vehicles)) # at least one!

# Find the NA
any(is.na(vehicles$id))
any(is.na(vehicles$make))

```

### Subsetting {-}

Recall that dataframes are filtered by row and/or column using this format: **`dataframe[rows,columns]`**. To get the third element of the second column, for example, you type `dataframe[3,2]`.

```{r,echo=TRUE,collapse=TRUE}
vehicles[3,2]
```

Note that the comma is necessary even if you do not want to specify columns. If you try to type this ...

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
vehicles[3]
```

...`R` will assume you are asking for the third column, not the third row.  

To filter a dataframe to multiple values, you can specify vectors for the `row` and `column`

```{r,echo=TRUE,collapse=TRUE}
vehicles[1:3,11:12] # can use colons
vehicles[1:3,c(1,11:12)]  # can use c()
```

Columns can also be called according to their names. Use the **`$`** sign to specify a column.

```{r,echo=TRUE,collapse=TRUE}
vehicles$hwy[1:5]
```

Note that when you use a `$`, you will not need to use a comma within your brackets. If you try to run this ...

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
vehicles$hwy[1:5,]
```

...`R` will throw a fit.  

Also recall that you can use logical tests, which return boolean values `TRUE` or `FALSE`, to filter dataframes to rows that meet certain conditions. For example, to filter to only the rows for cars with better than 100 mpg, you can use this syntax:

```{r,echo=TRUE,collapse=TRUE}
# Build your logical test
verdicts <- vehicles$hwy > 100

# Subset with booleans
vehicles[verdicts,2:3]
```

```{r,echo=TRUE,collapse=TRUE}
# Write this in one line to be more efficient:
vehicles[ vehicles$hwy > 100  , 2:3]
```

Recall that the logical test is returning a bunch of `TRUE`'s and `FALSE`'s, one for each row of `vehicles`. Only the `TRUE` rows will be returned.  


#### Exercise 2: Reading for errors {-}

What is wrong with these commands? Why will each of them throw an error if you run them, and how can you fix them?

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
# 1
vehicles[1,15,]

# 2
vecihles[1:5,]

# 3
vehicles$hwy[1:5,]

# 4
vehicles[1:5,1:13]
```

### Summarizing {-}

The same summary functions that you have used for vectors work for dataframes, such as:

```{r,echo=TRUE,collapse=TRUE}
min(vehicles$hwy)
```

```{r,echo=TRUE,collapse=TRUE}
max(vehicles$hwy)
```

```{r,echo=TRUE,collapse=TRUE}
mean(vehicles$cty)
```

```{r,echo=TRUE,collapse=TRUE}
sd(vehicles$cty)
```

```{r,echo=TRUE,collapse=TRUE}
str(vehicles$make)
```

```{r,echo=TRUE,collapse=TRUE}
class(vehicles$hwy)
```

You can also use the **`summary()`** function, which provides summary statistics for each column in your dataframe:

```{r,echo=TRUE,collapse=TRUE}
summary(vehicles)
```

The function **`unique()`** returns unique values within a column:

```{r,echo=TRUE,collapse=TRUE}
unique(vehicles$fuel)
```

Finally, the **`order()`** function helps you sort a dataframe according to the values in one of its columns.

```{r,echo=TRUE,collapse=TRUE}
 #sort dataframe by highway mileage
# only keep certain columns
vehicles_sorted <- vehicles[order(vehicles$hwy), 
                            c(2,3,4,10:12)] 
head(vehicles_sorted)
```

Reverse the order by wrapping **`rev()`** around the `order()` call:

```{r,echo=TRUE,collapse=TRUE}
vehicles_sorted <- vehicles[rev(order(vehicles$hwy)),
                            c(2,3,4,10:12)]
head(vehicles_sorted)
```


### Building dataframes {-}

As shown above, to create a new dataframe, use the `data.frame()` function.

```{r,collapse=TRUE}
my_vehicles <- data.frame(car=paste(vehicles$make,vehicles$model),
                          mgp_hwy=vehicles$hwy,
                          mpg_city=vehicles$cty)
my_vehicles[100:106,]
```

Note how the columns were named in the `data.frame()` call, and that each column is separated by a comma.

You can also stage an empty dataframe, which sounds useless but will become very useful as you start working with `for` loops and other higher-order `R` tools.  

```{r,echo=TRUE,collapse=TRUE}
df <- data.frame()
df
```

To coerce an object into a format that `R` interprets as a dataframe, use **`as.dataframe()`**: 

```{r,echo=TRUE,collapse=TRUE}
df <- as.data.frame(vehicles)
df[1:4,1:4]
```

You can bind multiple dataframes together using **`rbind()`**: 

```{r,echo=TRUE,collapse=TRUE}
df1 <- data.frame(name=c("Ben","Joe","Eric","Isabelle"),
                  instrument=c("Nose harp","Concertina","Ukelele","Drums"))
df1
```

```{r,echo=TRUE,collapse=TRUE}
df2 <- data.frame(name=c("Matthew"),
                  instrument=c("Washboard"))
rbind(df1,df2)
```

Note that to be combined, two dataframes have to have the exact same number of columns and the exact same column names.

The only exception to this is adding a dataframe with content an empty dataframe. That can work, and that will be helpful in the *`R` Toolbag* modules ahead.

```{r,echo=TRUE,collapse=TRUE}
df <- data.frame() # stage empty dataframe

df1 <- data.frame(name=c("Ben","Joe","Eric","Isabelle"),
                  instrument=c("Nose harp","Concertina","Ukelele","Drums"))

df <- rbind(df,df1)

df
```

You can also bind multiple dataframes together using **`cbind()`**: 
```{r,echo=TRUE,collapse=TRUE}
df1 <- data.frame(name=c("Ben","Joe","Eric","Isabelle"),
                  instrument=c("Nose harp","Concertina","Ukelele","Drums"))

df <- data.frame(age=c(33,35,35,20), home=c("Canada","Spain","USA","USA"))

df <- cbind(df,df1)

df
```

Note that to be combined, two dataframes have to have the exact same number of rows and the exact same column names.

### Exercise 3: Subsetting and filtering {-}

**A. Subset one field according to a logical test**  
With no more than two lines of code, get the number of Honda cars in the `vehicles` dataset.  

**B. Subset one field according to a logical test for a different field.**  
In a single line of code, show the mileages of all the Toyotas in the dataset.

**C. Subset a dataframe to a single subgroup**  
In a single line of code, determine how many differet car makes/models were produced in 1995.

**D. Get the mean value for a subgroup of data**  
What is the average city mileage for Subaru cars in the dataset?

**E. Subset a dataframe to only data from between two values**  
According to this dataset, how many different car makes/models have been produced with highway mileages between 30 and 40 mpg?

#### Review assignment {-}

1. Create a vector called `people` of 5 peoples names from the class.

2. Show with code how many people are in your vector

3. Create another vector called `height` which is the number of centimeters tall each of those 5 people are.

4. Combine these two vectors into a data frame.

