# Dataframe wrangling {#dplyr}

#### Learning goals {-}

* Understand the importance of *tidy* dataframes
* Understand what the `tidyverse` is and why it is awesome
* Feel comfortable working with dataframes using `dplyr` functions.


## The `dplyr` package {-}

Data scientists largely work in data frames and _do things_ to data. This is what the package `dplyr` is optimized for. It consists of a series of "verbs" which cover 95% of what you need to do for most basic data processing tasks. 

```{r,echo=TRUE,collapse=TRUE,eval=FALSE}
install.packages('dplyr')
```

```{r,echo=TRUE,message=FALSE}
library(dplyr)
```

The `dplyr` package contains a set of `verbs`: things you do to dataframes. Those verbs are:

- **`filter()`**  
- **`arrange()`**  
- **`select()`**   
- **`rename()`**  
- **`distinct()`**  
- **`mutate()`**   
- **`summarise()`**  
- **`sample_n()`** (and **`sample_frac()`**)  

## The `%>%` pipe {-}

**`%>%`** is a "pipe". It is a way to write code without so many parentheses. For example, what if I want to find the square root of the sum of a sample of two integers from a sequence of 10 to 20 by 2:

```{r,echo=TRUE,collapse=TRUE}
sqrt(sum(sample(seq(10, 20, 2), 2)))
```

The above could also be written as:

```{r,echo=TRUE,collapse=TRUE}
seq(10, 20, 2) %>% sample(2) %>% sum %>% sqrt
```

When you see the **`%>%`** pipe symbol, think of the word **"then"**. 

The above code could be read aloud like so: "First, get a sequence of every second numbers between 10 and 20. **Then**, take two random samples from that sequence. **Then**, sum those samples together. **Then**, take the square root of the sum." 

Using the `%>%` pipe framework, your code turns from a nonlinear series of parentheses and brackets to a linear progression of steps. 

Here's another example:

```{r,echo=TRUE,collapse=TRUE}
mean(sd(log(1:100)))
```

... could also be written as:

```{r,echo=TRUE,collapse=TRUE}
1:100 %>% log %>% sd %>% mean
```

## `dplyr` verbs {-}

### `filter()` {-}

The `filter()` function is used to subset a dataframe, retaining all rows that satisfy your conditions. To be retained, the row must produce a value of `TRUE` for all conditions. Note that when a condition evaluates to `NA` the row will be dropped, unlike base subsetting with `[ ... ]`.

```{r,echo=TRUE,collapse=TRUE}
people <- data.frame(who = c('Joe', 'Ben', 'Xing', 'Coloma'),
                    sex = c('Male', 'Male', 'Female', 'Female'),
                    age = c(35, 33, 32, 34))
people
```

```{r,echo=TRUE,collapse=TRUE}
people %>% filter(sex == 'Male')
```

```{r,echo=TRUE,collapse=TRUE}
people %>% filter(sex == 'Female')
```

### `arrange()` {-}

Arrange means putting things in order. That is, `arrange()` orders the rows of a data frame by the values of selected columns.

Unlike other dplyr verbs, arrange() largely ignores grouping; you need to explicitly mention grouping variables (or use .by_group = TRUE) in order to group by them, and functions of variables are evaluated once per data frame, not once per group. For example:

```{r,echo=TRUE,collapse=TRUE}
people %>% arrange(age)

people %>% arrange(sex)

people %>% arrange(who)

# Reverse the order
people %>% arrange(desc(age))

```

### `select()` {-}

Select (and optionally rename) variables in a data frame, using a concise mini-language that makes it easy to refer to variables based on their name (e.g. a:f selects all columns from a on the left to f on the right). You can also use predicate functions like is.numeric to select variables based on their properties.

```{r,echo=TRUE,collapse=TRUE}
people %>% select(age)
```

```{r,echo=TRUE,collapse=TRUE}
people %>% select(sex, age)
```

```{r,echo=TRUE,collapse=TRUE}
people %>% select(sex, years = age)
```


### `rename()` {-}

The function `rename()` changes the names of individual variables using new_name = old_name syntax; `rename_with()` renames columns using a function.

```{r,echo=TRUE,collapse=TRUE}
people %>% rename(gender = sex, years = age, first_name = who)
```


### `mutate()` {-}

The function `mutate()` adds new variables and preserves existing ones; `transmute()` adds new variables and drops existing ones. New variables overwrite existing variables of the same name. Variables can be removed by setting their value to NULL.

```{r,echo=TRUE,collapse=TRUE}
people %>% mutate(agein2020 = age - 1)
```

```{r,echo=TRUE,collapse=TRUE}
people %>% mutate(is_male = sex == 'Male')
```

```{r,echo=TRUE,collapse=TRUE}
people %>% mutate(who = NULL)
```

```{r,echo=TRUE,collapse=TRUE}
people %>% mutate(average_age = mean(age))
```

```{r,echo=TRUE,collapse=TRUE}
people %>% mutate(average_age = mean(age)) %>%
  mutate(diff_from_avg = age - average_age)
```


### `group_by()` {-}

Most data operations are done on groups defined by variables. The function `group_by()` takes an existing table and converts it into a grouped tbl where operations are performed "by group". `ungroup()` removes grouping.

```{r,echo=TRUE,collapse=TRUE}
people %>% 
  group_by(sex) %>%
  mutate(average_age_for_sex = mean(age)) 
```

```{r,echo=TRUE,collapse=TRUE}
people %>% 
  group_by(sex) %>%
  mutate(average_age_for_sex = mean(age)) %>%
  mutate(diff_from_avg_for_sex = age - average_age_for_sex)
```

### `summarize()` {-} 

`summarize()` or `summarize()` creates a new data frame. It will have one (or more) rows for each combination of grouping variables; if there are no grouping variables, the output will have a single row summarizing all observations in the input. It will contain one column for each grouping variable and one column for each of the summary statistics that you have specified.

```{r,echo=TRUE,collapse=TRUE}
people %>% 
  summarize(average_age = mean(age))
```

```{r,echo=TRUE,collapse=TRUE}
people %>% 
  summarize(average_age = mean(age),
            standard_dev_of_age = sd(age),
            oldest_age = max(age),
            youngest_age = min(age))
```

```{r,echo=TRUE,collapse=TRUE}
people %>% 
  group_by(sex) %>%
  summarise(avg_age = mean(age),
            oldest_age = max(age),
            total_years = sum(age)) 
```

```{r,echo=TRUE,collapse=TRUE}
people %>%
  group_by(sex) %>%
  summarise(sample_size = n())
```


#### Review exercise {-}


1. Run the below code.

```{r,echo=TRUE,collapse=TRUE}
library(dplyr)
library(babynames)
bn <- babynames
```

2. Create a dataframe named `turn_of_century. This should be data on baby names for the year 1900.

3. Create a dataframe named `boys`. This should be only males.

4. Create a dataframe named `moms_gen`. This should be females born in the year of birth of your mom.

5. Order `moms_gen` by `n`, in ascending order (ie, the least popular name at top). Look at the result; what is the least popular name among women the year your mom was born?

6. Reverse the order and save the result into an object named `moms_gen_ordered`.

7. Create an object named `boys2k`. This should be all males born in the year 2000.

8. Arrange `boys2k` from most to least popular. What was the most popular boys name in 2000?

9. What percentage of boys were named `Joseph` in 2000?

10. Were there more Jims or Matthews in 2020?

11. How many people were born with _your_ name in 2020?

