# Distributions & histograms 

#### Learning goals {-}

* How to go about exploring a dataset and understanding the distribution of your data  
* How to produce beautiful histograms and violin plots  

&nbsp;

Once you get your dataset into `R`, the first thing you will probably want to do is explore. Until you understand what your data look like, it is difficult to know how to analyze it. **Exploratory data analysis** is the process of becoming familiar with your data - its sample size, its distribution, its central tendencies, and its variation. This module shows you how. 

To practice exploring datasets, we will use one of *AirBnB* listings in Amsterdam ([go here](http://insideairbnb.com/get-the-data.html) and download `listings.csv.gz` under the Amsterdam section). 

```{r,echo=FALSE}
df <- read.csv("./data/airbnb-amsterdam.csv")
```

```{r,echo=TRUE,collapse=TRUE}
# Checkout what the data look like
# use t() to make it easier to read (t = transpose)
t(head(df,1))

# How much data do we have?
nrow(df)

```


## Exploring distributions {-}

One of the first steps in your data exploration will be to visualize your dataset. There are two main plot types for doing so: **histograms** and **violin plots**. 

### Histograms {-}

A histogram sorts your data into bins and counts the number of data points in each bin.  

Make histograms in `R` using the `hist()` function:  

```{r,echo=TRUE}
hist(df$availability_365)
```

This is a histogram of the number of nights the listings are available for reservation. In a quick glance, you can see that most listings are available for only a few nights, while a small portion have a wide variety of availabilities.   

Think about how much more helpul this image is than a simple sample mean. To illustrate that value, let's plot the mean onto this histogram. 

```{r,echo=TRUE}
hist(df$availability_365)
abline(v= mean(df$availability_365),col="red")
```

The average value of this dataset is nowhere near the most common values (1 - 20 nights), nor does it capture the fact that there are some listings available for the entire year.  **Punchline:** a lot of rich insight is lost when taking an average, which is why a histogram is more useful for displaying distributions than bar graphs (which only show the mean).  

#### Prettifying histograms {-}

Let's take a look at another variable: the price of lodging per night. 

```{r,echo=TRUE}
hist(df$price)
```

Not a very pretty plot. Since the range in prices is so wide, the default parameters for the `hist()` will need some tweaking. 

To make this histogram prettier, first let's manually set the number of "breaks", i.e., the cutoff values for each bin along the x-axis. 

```{r,echo=TRUE}
# Calculate breaks -- make each bin one day
breaks <- seq(0,max(df$price),by=5)

hist(df$price,
     breaks=breaks)
```

That's better. Let's zoom in on the most common values using `xlim()`. 

```{r,echo=TRUE}
hist(df$price,
     breaks=breaks,
     xlim=c(0,500))
```

Finally, let's improve the labels and colors. 

```{r,echo=TRUE}
par(mar=c(4.2,4.2,.5,.5))
hist(df$price,
     breaks=breaks,
     xlim=c(0,500),
     xlab="Listing price ($)",
     main=NULL,
     col=adjustcolor("darkblue",alpha.f=.6),
     border=adjustcolor("white",alpha.f=.1),
     )
```

Interesting spikes in the histogram. Why do you think those are occurring? That is a really interesting feature of the dataset that you might never have discovered if you had not plotted your histogram. 

#### Histograms in `ggplot2` {-}

To produce a histogram of the same data in `ggplot2`, use the following code: 
```{r,echo=TRUE}
library(ggplot2)
ggplot(df, aes(x=price)) + 
  xlim(0,500) +
  geom_histogram(binwidth=5,
                 fill=adjustcolor("darkblue",alpha.f=.6),
                 color=adjustcolor("white",alpha.f=.1)) 
```

`ggplot2` allows you to add some other nice features to histograms, such as a smoothed "density" line:

```{r,echo=TRUE,warning=FALSE}
ggplot(df, aes(x=price)) + 
  xlim(0,500) +
  geom_histogram(binwidth=5,
                 fill=adjustcolor("darkblue",alpha.f=.6),
                 color=adjustcolor("white",alpha.f=.1),
                 aes(y=..density..)) +
  geom_density(alpha=.1,lwd=1.5,col=adjustcolor("firebrick",alpha.f=.5))
```


### Violin Plots {-}

The second most valuable plot for exploring distributions is the **violin plot**. 

```{r,echo=TRUE}
ggplot(df, aes(x=room_type,
               y=price)) + 
  geom_violin()
```

Violin plots are like *vertical histograms*. Their width indicates where most of your data are clustered, and the shapes also take the shape of a violin -- hence the name. Their height shows you your data's range. Think of a violin plot as the beautiful child of bar graphs and histograms: the best of both worlds.  

A major advantage of violin plots is that they allow you two compare multiple distributions with ease. Let's adjust our `ylim()` to zoom in:

```{r,echo=TRUE,warning=FALSE}
ggplot(df, 
       aes(x=room_type,
           y=price)) + 
  ylim(0,500) +
  geom_violin()
```

If your research question were, "Which type of listing tends to be most affordable?", think about how easy that is to answer using a plot like this! 


## Descriptive statistics {-} 

In the [Module on Calling Functions](#calling_function), you have already seen some functions for summarizing the trends and variation in your data: 

- `mean()`
- `median()`
- `sd()` (standard deviation)
- `summary()` provides descriptive statistics for each column in a dataframe. 

Here are a few more:  

**`var()`** provides the sample variance, which is related to standard deviation:  

```{r,echo=TRUE,collapse=TRUE}
var(df$price)
```

**`MeanCI()`**, from the package `DescTools`, provides 95% confidence intervals for the most common data distributions. 

```{r,eval=FALSE,echo=TRUE,collapse=TRUE}
install.packages("DescTools")
```

```{r,echo=TRUE,collapse=TRUE}
library(DescTools)
MeanCI(df$price)
```

And here are some other useful functions that need not be introduced in detail now. You will encounter them soon.

- **`table()`** Counts the number of instances of each unique value in a vector.
- **`quantile()`** Provides the quantiles of a dataset.
- **`ecdf()`** Plots the empirical cumulative distribution function for a dataset.


#### Review exercise {-}

For this exercise, use this dataset on the sale prices of homes in the United States. You can download that dataset directly into `R` using the following code:

```{r,echo=TRUE,eval=FALSE}
library(readr)
df <- readr::read_csv("https://raw.githubusercontent.com/rashida048/Datasets/master/home_data.csv")
```

##### Task 1 {-}

Create a beautiful histogram of home prices included in this dataset. Only show houses that sold for less than $2 million USD.   

##### Task 2 {-}

Create a violin plot that shows the distribution of sale prices for homes with different numbers of bedrooms. 

##### Task 3 {-}

This task is a bit of review, and a bit of a preview to the next module.  

First, create a scatter plot that depicts the relationship between the square footage of a home's living space and its sale price. Do the same for the square footage of the home's lot. Which square footage appears to be a better predictor of sale price? 









