# (PART) Review exercises {-}


# A `dplyr` mystery

#### Group activity {-}

Using the `deaths` data, we're going to answer some questions together. Start with:

```{r,echo=TRUE,collapse=TRUE}
library(readr)
library(dplyr)
df <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/deaths.csv')
```

1. How many people are in the dataset.

2. Use `summarize` to count the number of men and women.  

3. Use `summarize` to count the number of people in each class.  

4. Use `summarize` to count the number of men and women in each class.  

5. Use `mutate` to create a variable called `died`. This should be a boolean based on the `Survived` column (in which 1 means the person survived, and 0 means the person died).

6. Use `mutate` to create a variable called `child`. This should be a boolean based on the `Age` column, indicating if someone was less than 18 years old.

7. Create a different dataframe for men vs. women. Name them accordingly.

8. Create a different dataframe for class 1, class 2, and class 3. Name them accordingly.

9. For each of the 5 datasets you've just created, what is the death rate?

10. For each of the 5 datasets, how many children died?

11. Now, using the _full_ dataset, calculate the child-specific death rate for each combination of class and sex (ie, "first class females", "third class males", etc.). 

12. What did you find? What might explain that?


# A `dplyr` survey


#### Team exercises {-}

Before getting started, start a new script (ctrl + shift + n), name it `survey_exercises.R`, and copy-paste the following code at the top.

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
# Load library
library(gsheet)
library(dplyr)

# Read in data
survey  <- gsheet::gsheet2tbl('https://docs.google.com/spreadsheets/d/1iVt9FX9J2iv3QFKBM7Gzb9dgva70XrW1lxMV4hpekeo/edit?resourcekey#gid=204634767')

# remake the names
names(survey) <- c('time', 'sex', 'age','sib', 'dad_mus', 'person_mus', 'joe_mus_is', 'eyesight', 'height', 'shoe_size', 'bday', 'money_or_love', 'rps_skill', 'num_pan', 'cats_dogs', 'name')

```


1) Create a dataframe called `old_people`. This should include only people older than 20. Write code to calculate the number rows in your new dataframe.

2) Create a dataframe called `captivated`. This should include all those people who find Joe's moustache to be "deeply captivating". Write code to calculate the number rows in your new dataframe.


3) Create a dataframe called `special_people`. This should be people who are taller than 175cm, prefer cats over dogs, and consider themselves to be average at rock, paper, scissors.

4) In the full dataset (survey), do more people like cats or dogs? What about among "special" people?

5) Create a new variable in `survey` called "std_shoes" that standardizes shoe sizes by converting men's shoe size to women's (There is an approximate 1.5 size difference between Men's and Women's sizing (e.g., a men's size 7 is roughly equivalent to a women's size 8.5)

6) Get the avg shoe size by sex (Male, Female, Prefer not to say) 

7) Get Average age, height, & number of siblings, by the sex

8) Do people that have ever had a mustache think there will be more pandemics on average than those who have never had a mustache?

9) Do people that prefer cats have smaller feet on average than those who prefer dogs?

10) Is eyesight associated with moustache perception?

11) What percentage of people think they are better than average at rock scissors paper?

12) What percentage of men and women think they are better than average at rock scissors paper?

13) How many people think money matters more than love?

14) Create a dataframe, grouped by whether or not people's dads had moustaches, with variables showing each of the following: the maximum age, maximum height,minimum number of pandemics, and average number of siblings

15) What percentage of men have terrible eyesight?

16) How many women have a shoe size of 9 or more?

17) Create a variable in the `survey` dataset named `days_old`. Use the `bday` variable and subtract it from `Sys.Date()`.

18) What is the standard deviation of age?

19) Create a one-column dataset which contains the name(s) of the person(s) with the most number of siblings (hint: use the following dplyr verbs in this order: `filter`, `select`).

20) Tell me something interesting about this dataset. 

```{r, eval = FALSE, echo = FALSE}

# 10) Create a bar chart that compares the average age of those who prefer cats vs dogs
# 
# 11) Create a scatter plot that shows the relationship between age and height (make it look nice!)
# 
# 12) Create a scatter plot that shows the relationship between height and shoe size 
# 
# 13) Create a bar chart that shows the average number of siblings for dad's mustache status.
# 
# 14) Color the barchart above and add some degree of transparency in the color. Add a title.
# 
# 15) Create a bar chart that shows the average shoe size by sex and cat/dog preference. 
# 
# 

```
