# (PART) Review exercises {-}


```{r, results = 'asis', eval = ifelse(exists('is_teacher'),is_teacher,TRUE),echo=FALSE}
source('teacher_tip.R')

tip <- 'This review exercise is best done immediately after the module on **Dataframe wrangling**.'

teacher_tip(tip)
```

# A `dplyr` mystery

Use the `dplyr` verbs to answer these questions.  

First, download the following mystery dataset by running this code.  

```{r,echo=TRUE,collapse=TRUE, message = FALSE}
library(readr)
library(dplyr)
df <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/deaths.csv')
```

**1.** Review the dataset and try to figure out which each row represents. "Each row is a ____."  

**2.** How many people are in the dataset?  

**3.** Use `summarize()` to count the number of men and women.  

**4.** Use `summarize()` to count the number of people in each class.   

**5.** Use `summarize()` to count the number of men and women in each class.   

**6.** What is the average age of men in the dataset?

**7.** What is the average age of women in the dataset? 

**8.** Use `mutate` to create a variable called `died`. This should be a boolean based on the `Survived` column (in which 1 means the person survived, and 0 means the person died).  

**9.** Use `mutate` to create a variable called `child`. This should be a boolean based on the `Age` column, indicating if someone was less than 18 years old.  

**10.** Create a different dataframe for men vs. women. Name them accordingly.  

**11.** Create a different dataframe for class 1, class 2, and class 3. Name them accordingly.  

**12.** For each of the 5 datasets you've just created, what is the death rate?  

**13.** For each of the 5 datasets, how many children died?  

**14.** Now, using the _full_ original dataset, calculate the child-specific death rate for each combination of class and sex (ie, "first class females", "third class males", etc.).  

**15.** What did you find? What might explain that?  

**16.** What is the average age of men and women, separately, in each class?

**17.** What do you think this dataset represents? Where does it come from?  


# A `dplyr` survey

```{r, results = 'asis', eval = ifelse(exists('is_teacher'),is_teacher,TRUE),echo=FALSE}
source('teacher_tip.R')

tip <- 'This review exercise is best done immediately after the module on **Dataframe wrangling**.'

teacher_tip(tip)
```

Use the `dplyr` verbs to answer these questions.  

First, download the results of a recent survey by running this code.  

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
# Load library
library(gsheet)
library(dplyr)

# Read in data
survey  <- gsheet::gsheet2tbl('https://docs.google.com/spreadsheets/d/1iVt9FX9J2iv3QFKBM7Gzb9dgva70XrW1lxMV4hpekeo/edit?resourcekey#gid=204634767')

# remake the names
names(survey) <- c('time', 'sex', 'age','sib', 'dad_mus', 'person_mus', 'joe_mus_is', 'eyesight', 'height', 'shoe_size', 'bday', 'money_or_love', 'rps_skill', 'num_pan', 'cats_dogs', 'name')

```

**1.** Review the `survey` dataset and try to figure out which each row represents. "Each row is a ____."  

**2.** How many respondents are in the dataset?  

**3.** Create a dataframe called `old_people`. This should include only people older than 20. Write code to calculate the number rows in your new dataframe.

**4.** Create a dataframe called `captivated`. This should include all those people who find Joe's moustache to be "deeply captivating". Write code to calculate the number rows in your new dataframe.

**5.** Create a dataframe called `special_people`. This should be people who are taller than 175cm, prefer cats over dogs, and consider themselves to be average at rock, paper, scissors.

**6.** In the full dataset (survey), do more people like cats or dogs? What about among "special" people?

**7.** Create a new variable in `survey` called "std_shoes" that standardizes shoe sizes by converting men's shoe size to women's (There is an approximate 1.5 size difference between Men's and Women's sizing (e.g., a men's size 7 is roughly equivalent to a women's size 8.5)

**8.** Get the avg shoe size by sex (Male, Female, Prefer not to say) 

**9.** Get Average age, height, & number of siblings, by the sex

**10.** Do people that have ever had a mustache think there will be more pandemics on average than those who have never had a mustache?

**11.** Do people that prefer cats have smaller feet on average than those who prefer dogs?

**12.** Is eyesight associated with moustache perception?

**13.** What percentage of people think they are better than average at rock scissors paper?

**14.** What percentage of men and women think they are better than average at rock scissors paper?

**15.** How many people think money matters more than love?

**16.** Create a dataframe, grouped by whether or not people's dads had moustaches, with variables showing each of the following: the maximum age, maximum height,minimum number of pandemics, and average number of siblings

**17.** What percentage of men have terrible eyesight?

**18.** How many women have a shoe size of 9 or more?

**19.** Create a variable in the `survey` dataset named `days_old`. Use the `bday` variable and subtract it from `Sys.Date()`.

**20.** What is the standard deviation of age?

**21.** Create a one-column dataset which contains the name(s) of the person(s) with the most number of siblings (hint: use the following dplyr verbs in this order: `filter`, `select`).

**22.** Create a bar chart that compares the average age of those who prefer cats vs dogs

**23.** Create a scatter plot that shows the relationship between age and height (make it look nice!)

**24.** Create a scatter plot that shows the relationship between height and shoe size 

**25.** Create a bar chart that shows the average number of siblings for dad's mustache status.

**26.** Color the barchart above and add some degree of transparency in the color. Add a title.

**27.** Create a bar chart that shows the average shoe size by sex and cat/dog preference. 

