# Working with dates & times {#dates}


```{r,echo=FALSE,eval=FALSE}
is_teacher <- FALSE
```

#### Learning goals {-}

* Be able to read dates, and convert objects to dates
* Be able to convert dates, extract useful information, and modify them
* Use date times
* Gain familiarity with the lubridate package

```{r, results = 'asis', eval = is_teacher}
teacher_content('Here is some teacher content.')
```

## First steps  {-}

Hadley Wickham's [tutorial on dates](https://r4ds.had.co.nz/dates-and-times.html) starts with 3 simple questions:


-   Does every year have 365 days?
-   Does every day have 24 hours?
-   Does every minute have 60 seconds?

"I'm sure you know that not every year has 365 days, but do you know the full rule for determining if a year is a leap year?
(It has three parts.) You might have remembered that many parts of the world use daylight savings time (DST), so that some days have 23 hours, and others have 25.
You might not have known that some minutes have 61 seconds because every now and then leap seconds are added because the Earth's rotation is gradually slowing down.

Dates and times are hard because they have to reconcile two physical phenomena (the rotation of the Earth and its orbit around the sun) with a whole raft of geopolitical phenomena including months, time zones, and DST.
This chapter won't teach you every last detail about dates and times, but it will give you a solid grounding of practical skills that will help you with common data analysis challenges."

## Lubridate  {-}

Let's install the `lubridate` package.

## Getting familiar with the `date` type  {-}

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
library(lubridate)
today <- today()
my_birthday <- '1985-11-07'
str(today)
str(my_birthday)

# This causes an errors
# today - my_birthday

# This does not
my_birthday <- as_date(my_birthday)
today - my_birthday
```

## Converting dates from strings  {-}

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```

What does the `tzone` argument to `today()` do? Why is it important?

Use the appropriate lubridate function to parse each of the following dates:

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
```

## Getting familiar with the `datetime` type  {-}

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
n <- now() 
n + seconds(1)
n - hours(5)
hour(n)
minute(n)
as_date(n)
later <- now()
```


## Extracting components from date and datetime objects  {-}


```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
datetime <- ymd_hms("2016-07-08 12:34:56")
year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
weekdays(datetime)
```


## Cough data exercise  {-}


```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
coughs <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/coughs.csv')
```

1. Create a `dow` (day of week) column.

2. Create a `date` (without time) column.

3. How many coughs happened each day?

4. Create a chart of coughs by day.

5. Look up `floor_date`. Use it to get the number of coughs by date-hour.

6. Create an `hour` variable.

7. Use the `hour` variable to create a `night_day` column indicating whether the cough was occurring at night or day.

8. Does Galileo cough more at night or day?


## Weather data exercise {-}

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
library(weatherr)
library(tidygeocoder)
sewanee_location <- geo(address = '735 University Avenue, Sewanee TN')
fc <- locationforecast(lon = sewanee_location$long,
                       lat = sewanee_location$lat)

```

1. Plot the forecasted temperature

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
ggplot(data = fc,
       aes(x = time,
           y = temperature)) +
  geom_point()
```

2. Get average cloudiness by day.

3. Plot average cloudiness by day.

4. Plot humidity.

5. Get the maximum temperature by hour (without date). Plot it.

6. Make a variable called `date`. This should just be the date (without the time)

7. Plot wind direction over time, with the line color showing the date.

8. Create a variable called `perfect`. This should include conditions indicating whether the weather in that moment is perfect.

9. How much of the next few days is going to be perfect?

10. Create a variable called `hour`. What is the best time of day (across all days) to fly a kite in the next few days?

11. Use `date` to get the average windspeed per day. What is the best day to fly a kite this week?

12. Get the average temperature by `hour`. What is, on average, the coldest moment of the day? And the hottest?



