# Working with text

## Learning goals {-}

* Item 1  
* Item 2  
* Item 3 


## Basics {-}

### The `stringr` package {-}

```{r,echo=TRUE,eval=FALSE}
install.packages("stringr")
library(stringr)
```

## Common tools {-}

**`as.character()`**: converts a non-character object into a character string. 

```{r,echo=TRUE}
x <- c(1,2,3)
as.character(x)

x <- as.factor(c("group1","group2","group3"))
as.character(x)
```

This can be particularly useful if you are trying to avoid problems caused by factors. One common issue occurs when `R` mistakes a set of numbers as a set of factors. Using `as.character()` can set things right:

```{r,echo=TRUE}
x <- as.factor(c(18,19,20))
x

# If you try to convert straight to numeric, it does not work:
as.numeric(x)

# So convert to character first
as.numeric(as.character(x))
```

**`paste()`** and **`paste0()`**: description:

```{r,echo=TRUE}

```

**`tolower()`** and **`toupper()`**: description:

```{r,echo=TRUE}

```

**`nchar()`**: description:

```{r,echo=TRUE}

```

**`substr()`**: description:

```{r,echo=TRUE}

```

**`grep()`**: description:

```{r,echo=TRUE}

```

**`gsub()`**: description:

```{r,echo=TRUE}

```

**`stringr::str_pad()`**: description:

```{r,echo=TRUE}

```

**`str_split()`**: description:

```{r,echo=TRUE}

```

## Review assignment {-}

```{r,echo=TRUE}

```

