# Trump tweets {#trump}

#### Exercises (Trump tweets)  {-}

Let's run the below to get started.

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}

library(dplyr)
library(readr)
library(tidytext)
trump <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/trumptweets.csv')
stop_words <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/stop_words')

```

1. In the current format, one row of data is equal to one <what>?

2. Create a variable called `line`. This should be 1, 2, 3, 4, etc.

3. Create a variable called `text`. This should be an exact copy of `content`.

4. Use the `unnest_tokens` function to reshape the data for better text processing.

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
simple <- trump %>%
  select(-mentions, -hashtags, -geo, -content, -link, -id) %>%
  unnest_tokens(word, text)
```

5. What format is the data in now (ie, one row is equal to <what>)?

6. Take a minute to read about the `tidytext` package at https://www.tidytextmining.com/tidytext.html.

7. What is the most common word used by Trump?

8. Use `substr` to create a `year` variable.

9. What is the most common word used by Trump each year?

10. Create a variable named `month` using `substr`.

11. What is the most common word used by Trump each month?

12. Create a dataframe with one word per row, and a column called `freq` saying how many times that word was used.

13. Load up the `wordcloud` library.

14. Subset the dataframe created in number 12 to only include the top 100 words.

15. Create a wordcloud of Trump's top 100 words.

16. Are you ready to do some sentiment analysis? Great.

17. Create a dataframe named `sentiments` by running the following: `sentiments <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/sentiments.csv')`

```{r, eval = FALSE, echo = FALSE}
sentiments <- get_sentiments("nrc")
uns <- sort(unique(sentiments$sentiment))
sentiments <- sentiments %>%
  group_by(word) %>%
  summarise(anger = length(which(sentiment == 'anger')),
            anticipation = length(which(sentiment == 'anticipation')),
            disgust = length(which(sentiment == 'disgust')),
            fear = length(which(sentiment == 'fear')),
            joy = length(which(sentiment == 'joy')),
            negative = length(which(sentiment == 'negative')),
            positive = length(which(sentiment == 'positive')),
            sadness = length(which(sentiment == 'sadness')),
            surprise = length(which(sentiment == 'surprise')),
            trust = length(which(sentiment == 'trust')))
write_csv(out, 'data/sentiments.csv')
```

18. What is the `sentiments` dataset?

19. Create another dataset named `polarity` by running the following: `polarity <- get_sentiments("afinn")`

20. Use `left_join` to combine polarity and sentiments into one dataset named `emotions`.


```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
emotions <- left_join(sentiments, polarity) %>% filter(!duplicated(word))
```

21. Use `left_join` to combine the `trump` data and the `emotions` data.

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
simple <- left_join(simple, emotions)
```

22. Have a look at the `simple` (Trump) data. What do you see?

23. Get an overall polarity score (using the `value` variable) for the entire dataset. Is it positive or negative?

24. How many words were emotionally associated with "anger" in 2015?

25. What percentage of words were associated with "fear" by year?

26. What is the average sentiment polarity by year?

27. What is Trump's most positive tweet?

28. What month was Trump's most negative month?

29. What percentage of Trump tweets have more sadness than joy by year/month?

30. Read in data on full moons by running the following: `moon <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/full-moon.csv')`

31. Create a `date` column with a correctly formatted date.

32. What day of the week has the most full moons?

33. Use `left_join` to bring the moon data into the Trump data.

34. Does Trump have more negative emotions on full moon days?

35. Read in "stop words" by running the following: `sw <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/stopwords.csv')`

36. Join the `sw` data to the `simple` data, and remove the stop words.

37. Create a new word cloud.

38. Do a new analysis of sentimentality

39. Which day of the week does Trump have the most
