# Cleaning messy data


#### Learning goals {-}

* This is a review exercise: learning the skills introduced in the previous modules by applying them to a universal data science scenario: cleaning up messy data.  

```{r, echo=FALSE, eval=FALSE,collapse=TRUE}

mr <- read.csv("./data/whale-dives-messy.csv")
head(mr)

yyyy <- substr(mr$id,1,4) ; yyyy
mm <- substr(mr$id,5,6) ; mm
dd <- substr(mr$id,7,8) ; dd
sit <- substr(mr$id,9,11) ; sit

yyyy
yyyy[grep(5,yyyy)] <- "15"
yyyy
mr$YEAR <- yyyy

mr$Month <- as.numeric(mm)
mr$Day <- as.numeric(dd)

head(mr)

mr$sit <- paste0("Sighting ",sit) 
mr$id <- NULL

head(mr)

mr <- mr[sample(1:nrow(mr),size=nrow(mr),replace=TRUE),]
head(mr)

write.csv(mr,file="./data/whales-dives-messy.csv",quote=FALSE,row.names=FALSE)

```

### Your mission {-}


In the module on joining datasets, we introduced a dataset of whale diving behaviour: 

`r xfun::embed_file("./data/whales-dives.csv",text="whales-dives.csv")`  


```{r, echo=TRUE}
dives <- read.csv("./data/whales-dives.csv")
```

```{r, echo=TRUE,collapse=TRUE}
head(dives)
```

This dataframe is nice and tidy. Here are a few of its many tidy features: 

- Each row is a single observation.   

- There is not a single missing value anywhere in the dataset.  

- The rows are organized from earliest to most recent, based on the data embedded in the `sit` column.   

- Categorical columns have standardized formatting. In the `species` column, there are two levels: `HW` (humpback whale) and `FW` (fin whale). In the `behavior` column, there are also two levels: `FEED` and `OTHER`.  

But this dataset was not always so pretty. Here is the link to the **original** data file: 

`r xfun::embed_file("./data/whales-dives-messy.csv",text="whales-dives-messy.csv")`  


```{r, echo=FALSE,eval=FALSE}
messy <- read.csv("./data/whales-dives-messy.csv")
```

**Your task in this review exercise is to write a script that carries out the necessary data cleaning steps to get this dataset from its original form to its tidy form.** 

Test your work along the way, then demonstrate its completion, using the `identical()` function. If your `my_dives` version of the dataset is identical to the `dives` data above, the following logical test will be true: 

```{r, echo=TRUE,collapse=TRUE,eval=FALSE}
identical(dives,my_dives)
```

Enjoy!  





