# `R Markdown` 

#### Learning goals {-}

- How to create `RMarkdown` documents  
- Why they are awesome  
- Why they are essential to truly *reproducible* reporting.  

&nbsp;  

`R Markdown` is the combination of `R` and `Markdown`: you create a `Markdown` document that has `R` built-in to it: the two languages are woven, or "knitted", together.  

## Do first, explain later {-}

Let's not talk about it. Let's just do it.

#### Start an `RMarkdown` report {-}

**1.** Open Rstudio.  
**2.** Click `File` -> `New File` -> `R Markdown`  
**3.** Fill out "Title", "Author", and click "Set to HTML"  
**4.** Type  `Ctrl + S` to save your file in your working directory.   

#### Compile your report {-}

**5.** Have a look at the document. What does each section mean?  
**6.** Click the *"Knit"* button at the top.  
**7.** Cool, right? 

But also kind of boring. Let's make some changes to this report.  

#### Modify the template {-}

**8.** Let's replace the "setup chunk" with the below.  

```{r, include=FALSE, echo = TRUE, eval = TRUE}
# Basic knitr options
library(knitr)
opts_chunk$set(comment = NA, 
               echo = TRUE, 
               warning = FALSE, 
               message = FALSE, 
               error = TRUE)
```

**9.** Then, let's change the title to some interesting question.   

**10.** Now, let's delete all of the other stuff in the template so far (beginning at "R Markdown").   

**11.** Put in some section headers using `Markdown`. Make the following sections: 

```
# Introduction

# Methods

# Results

# Discussion

# Bibliography
```

**12.** Write a two-sentence introduction to your made-up research topic.  

**13.** `Knit` again, to make sure that we did not break anything.  (You should `knit` often!)  

Good? Great.  

## Add an `R` chunk {-} 

In your *Methods* section, let's create an `R` chunk in which we can generate some fake data.  

To make an `R` chunk, type this:

```` 
```{r}
# code goes here
```
````
Those three marks represent the start and end of a chunk of code.  

Now add this code into your chunk: 

```{r,echo=TRUE,eval=TRUE}
x <- 1:1000
y <- sqrt(x) + rnorm(1000,0,sd=3)
plot(y ~ x, cex=.7)
```

Know, re-`knit` your doc. Everything work? 

### Chunk options {-}  

#### Hide a chunk {-} 

To run the code within an `R` chunk but hide the code in your final output, change the bracketed code at the top of your chunk to read `{r, echo = FALSE}`.

#### Don't evaluate a chunk {-}

To skip over a chunk without evaluating it, change your brackets to look like this: `{r, eval = FALSE}`.  

What if you want to hide a chunk *and* skip over it? `{r, echo = FALSE, eval = FALSE}`. This is a nice way to store code scraps that you don't want to run but aren't ready to get rid of.  

#### Change figure size {-}  

To change the dimensions of a plot output, change your brackets to look like this:
`{r, fig.height = 3, fig.width = 7}`

Doing so produces a plot like this: 

```{r,fig.height=3, fig.width=7, echo=FALSE}
plot(y ~ x, cex=.7)
```

#### Add figure caption {-} 

To add a figure caption, change your brackets like so: `{r, fig.cap = "A squished plot."}`  
 
```{r,fig.height=3, fig.width=7, fig.cap="A squished plot.", echo=FALSE}
plot(y ~ x, cex=.7)
```

## In-line `R` code {-} 

In your *Results* section, add this line:

> Our analysis was based on  `` `r knitr::inline_expr("length(x)")` `` observations.  

Know, re-`knit` your doc. What was printed before the word "observations"? 

&nbsp;  

**This is why `RMarkdown` is reproducible.** Think of how a typical research report works: you get your results from
`R` or whatever program you are using, then you manually write those results numbers into a *Word* document. What is reproducible about that? 

With `RMarkdown`, you can let the data underlying your analysis *literally speak for themselves*.  


## Tables {-}  

To make a pretty table out of a dataframe, create a chunk that looks like this:  

```{r, results='as is'}
library(knitr)
df <- data.frame(x,y)
kable(head(df), caption = "A pretty sweet table.")
```


## Bibliographies  {-}

- In `RStudio`, create a new Text File and save it as `bibliography.bib` in the same location as your `RMarkdown` file.   

- Look up [this paper](https://www.researchgate.net/publication/260642613_Statistical_exponential_formulas_for_homogeneous_diffusion)  

- Copy the `DOI`.  

- Go to [https://doi2bib.org/](https://doi2bib.org/)  

- Copy the `bibtex` format   

- Paste into your `.bib` file.  

Note that the top line of this `bib` entry has the phrase `Rudd2014`. This is known as the citation's "key" -- you can use this phrase to reference this source within your `Markdown` doc.  

- Back in your `RMarkdown` doc, in your *Discussion* section, write some line about a paper.

- At the end of that sentence, add this text: `([@Rudd2014])`.  

- Add `bibliography: bibliography.bib` to your "`yaml`", the header information at the very top of your `RMarkdown` file. Change the `yaml` so that it looks like this:  

```
---
output: html_document
bibliography: bibliography.bib
---
```

- Re-`knit` your doc to see what happens. Was the citation automatically generated?  


#### Exercises {-}

Before getting started, think of a testable hypothesis which can be answered via a survey of your friends and classmates. For example: "woman have more dreams than men", or "people born abroad have shorter last names than people born in the USA", or "shoe size is associated with GPA".  

**1.** Create a repository on github called `report`.  

**2.**  Clone that repository to your local machine.  

**3.**  `cd` into that repository.  

**4.**  In that repo, create a new `Rmarkdown` document. Name it `paper.Rmd`. 

**5.** Replace the setup chunk with the one used in the in-class exercise (above).

**6.** Create a *GoogleForms* survey with at least 5 interesting questions on a topic which interests you (one topic, five questions).

**7.** Get at least 5 people to fill out your survey. 

&nbsp;  

**8.** In your *GoogleForms* settings, create a *GoogleSheet* of your results.  

**9.** In the sharing settings of your *GoogleSheet*, set to "public".  

**10.** Copy the URL of your *GoogleSheet*.  

**11.** Create an R chunk in your `paper.Rmd`.  

**12.** Use the `gsheet` package to read your survey results as a dataframe.

&nbsp;  

**13.** After reading in the data, create the following sections in your report: `Introduction`, `Methods`, `Results`, `Conclusion`. 

**14.** Spend 5-15 minutes on the internet reading about your topic / hypotheses. 

**15.**  Find 2-5 authors/articles of interest related to your topic. Cite them in the *Introduction*.

**16.**  Describe in plain English your approach to testing your hypothesis in the *Methods* section.  

**17.** In your *Methods* section, provide a URL link to your survey.   

&nbsp;  

**18.**  After the *Methods* section, write the `R` code (in an `echo`ed chunk) which tests your hypothesis.

**19.**  In the *Results* section, write a description of both (a) your sample and (b) your results. Include at least two charts. Include at least one table.

**20.** In the *Conclusion* section, write interesting reflection on your data.

&nbsp;  

**21.** Got here? Cool, you're fast. Now configure your computer to generate pdfs via Rmarkdown: [mac](https://medium.com/@sorenlind/create-pdf-reports-using-r-r-markdown-latex-and-knitr-on-macos-high-sierra-e7b5705c9fd) / [windows](https://medium.com/@sorenlind/create-pdf-reports-using-r-r-markdown-latex-and-knitr-on-windows-10-952b0c48bfa9)

**22.** Make your paper in pdf.  

**23.** Check out these [theme options for `Rmarkdown` docs.](https://www.datadreaming.org/post/r-markdown-theme-gallery/)

**24.** Push your code to git

**25.** Check out how to [make presentations in `Rmarkdown`.](https://rmarkdown.rstudio.com/lesson-11.html) Make one.  

**26.**  Check out how to [make dashboards in `Rmarkdown`.](https://rmarkdown.rstudio.com/lesson-12.html) Make one.

**27.** Change your citation style. To do so, [use a new csl from this site.](https://github.com/citation-style-language/styles)

