# Interactive maps

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}

```

#### Learning goals {-}

- Learn how to create interactive, *beautiful* maps with `leaflet`. 

- Understand why `leaflet` is so awesome and valuable in *reproducible* data science.  


## Mapping with `leaflet` {-}  

If you know about the right packages, making really nice maps in `R` is actually pretty easy. `R`'s spatial visualization tool set is one of the things that elevates it above the other open-source coding tools out there. 

Being able to make a map programmatically is a powerful skill. And we can do it quickly and easily in `R` thanks to the package `leaflet`.  

Let's load it up:   

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
install.packages('leaflet')
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE, message=FALSE}
library(leaflet)

# Some other packages too:
library(dplyr)
library(ggplot2)
```

To practice making a map, let's focus in on the location of Sewanee: The University of the South, in Tennessee, USA.  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
sewanee <- data.frame(longitude=-85.9211, 
                      latitude=35.2031)
```

To begin a `leaflet` map, simply type `leaflet()`. This creates a blank canvas. 

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet()
```

&nbsp;  

Now, add the background map. To do so, `leaflet` uses "tiles". Tiles are basically images -- little pictures stitched together to give you a sense of a continuous map.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% addTiles() 
```

&nbsp;  

Try zooming into this map to get a sense of how powerful these interactive `leaflet` maps can be.  

To get the map focused on a specific area of interest, bring in "markers" based on your data:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addTiles() %>% 
    addMarkers(data=sewanee)
```

&nbsp;  

Ta-dah! There's your map. Try zooming in & out again.  

### Stylizing your `leaflet` map {-}  

How about satellite imagery instead?  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addProviderTiles(providers$Esri.WorldImagery) %>% 
    addMarkers(data=sewanee)
```

&nbsp;  

How about topography instead?  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addProviderTiles(providers$Esri.WorldTopoMap) %>% 
    addMarkers(data=sewanee)
```

&nbsp;  

The `leaflet` package has plenty of other tile styles available through the `addProviderTiles()` function. To preview all the options, check out [this site.](https://leaflet-extras.github.io/leaflet-providers/preview/)   

### Working with markers in `leaflet` {-}  

To change the markers from a "dart" to a circle, use `addCircleMarkers()` instead: 

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addProviderTiles(providers$Esri.WorldTopoMap) %>% 
    addCircleMarkers(data=sewanee, 
                     color = "firebrick",
                     radius = 15)
```

&nbsp;  

To add 'pop up' information when you click on your marker, use the `popup` input:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addProviderTiles(providers$Esri.WorldTopoMap) %>% 
    addCircleMarkers(data=sewanee, 
                     color = "firebrick",
                     radius = 15,
                     popup = paste0('<b>YSR, yall.</b> <br>', 
                                    'Lat: ',sewanee$longitude,'<br>',
                                    'Lon: ',sewanee$latitude,'\n'))
```

&nbsp;  

Click on the marker and see how it looks! Note that we used a bit of `HTML` to make the text look prettier.  

### Mapping marine areas {-}  

Note that `leaflet` is also useful in mapping marine areas.  

For example, map the bathymetry of the northeast Pacific basin:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addProviderTiles(providers$Esri.OceanBasemap) %>%
    fitBounds(lng1 = -165,
              lng2 = -100,
              lat1 = 10,
              lat2 = 50)
```

&nbsp;  

Or get satellite imagery for an island off the coast of Georgia, USA:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addProviderTiles(providers$Esri.WorldImagery) %>% 
    fitBounds(lng1 = -81.2,
              lng2 = -81.1,
              lat1 = 31.54,
              lat2 = 31.7)
```


## Geocoding {-}  

To jumpstart your mapping skills in `R`, the second package you need to know about is `tidygeocoder`. This package helps you **"geocode"** a mailing address. Geocoding means providing the latitude and longitude for an address, which allows you to find it on a map.  

First, install & load the package:  

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
install.packages('tidygeocoder')
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
library(tidygeocoder)
```

Second, create a dataframe for the address(es) you wish to map:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
addresses <- data.frame(name= 'White House', 
                        addr = "1600 Pennsylvania Ave NW, Washington, DC")
```

Now, geocode:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE, message=FALSE}
lat_longs <- addresses %>%
    geocode(addr, method = 'osm', lat = latitude , long = longitude)
```

Check out the result:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
lat_longs
```

And check out your map:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE, message=FALSE}
leaflet(data=lat_longs) %>% 
    addProviderTiles(providers$Stamen.Watercolor) %>% 
    addMarkers()
```

&nbsp;  

You can also **reverse geocode**, i.e., get the mailing address for a lat-long coordinate.  

Say these are your coordinates of interest:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
lat_longs <- data.frame(latitude=48.8584, 
                        longitude=2.2945)
```

Now, reverse-geocode:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
reverse <- lat_longs %>%
    reverse_geocode(lat = latitude, 
                    long = longitude)
```

Check out the result:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
reverse
```

Now, add the mailing address as a `popup` in your marker:  

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
leaflet() %>% 
    addProviderTiles(providers$HikeBike.HikeBike) %>% 
    addCircleMarkers(data=lat_longs, 
                     color = "firebrick",
                     radius = 15,
                     popup = gsub(",",",<br>", reverse$address))
```


## Final thoughts {-}  

Maps, even simple ones, are amazingly effective data visualizations. The density and layers of information contained within a map are incredible!  

All of those virtues are compounded with an interactie map. At each zoom level, you gain a new perspective.  Like `shiny` apps, `leaflet` maps put the viewer in control of how they explore and understand the data.  



### Exercises {-}  

**1.** Create a new `Rmarkdown` document.

**2.** Create a code chunk. In this chunk, read in some data on "conflicts". To read in the data, run the below code.  

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
download.file('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/conflicts.RData',
              destfile = 'conflicts.RData')

load('conflicts.RData')
```

This data comes from [https://ucdp.uu.se/encyclopedia](https://ucdp.uu.se/encyclopedia). Take a minute or two to look at the website. 


**3.** Have a look in the data. Which fields are likely to be geographic?

**4.** Make a simple x-y plot using geographic fields.

```{r,echo=FALSE,collapse=TRUE, eval = FALSE}
ggplot(data = conflicts,
       aes(x = longitude,
           y = latitude)) +
  geom_point()
```

**5.** Create an object named `conflicts_afg`. This should be a plot of conflicts in Afghanistan.

**6.** Color the points by year.

**7.** Instead of year, color the points by `deaths_civilians`.

**8.** Color the points by date, but make point size reflect `deaths_civilians`.

**9.** Create a leaflet map of conflicts in a country of your choice.

**10.** Use `addTiles`.

**11.** Use `addProviderTiles` to make your map a satellite map.

**12.** Explore other tiles

**13.** Add pop-ups to your maps by using the `popup` argument within `addMarkers`.

**14.** Add `clusterOptions = markerClusterOptions()` to make your points clustered.

**15.** Replace your markers with "circle markers".

**16.** Create a shiny app wherein the user selects a country and time frame, and the app shows both (a) an interactive map and (b) a plot of the number of conflicts by year for that country. 

