# Sentiment analysis with Harry Potter


&nbsp;  

**1.** Start a new R file Name it "harry.R".

**2.** Set up your work space by loading the `ggplot2`, `dplyr`, `tidytext`, `gsheet`, `wordcloud2`, `sentimentr`, and `lubridate` packages.

**3.** Read in your Harry Potter data by running the following:

```{r, eval = FALSE}
hp <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/harrypotter.csv')
```

**4.** Take a peek at the first few rows of the data. What is the unit of observation?

**5.** Make all of the `text` column be lower case.

**6.** Make all of the `text` column be upper case.

**7.** Use `unnest_tokens` to create a dataframe with one row per word.

**8.** Create a variable named `word_length` with the number of characters in each word.

**9.** Make a histogram of word length.

**10.** Make a density chart of word length.

**11.** Make the density chart of word length have a different `fill` for each chapter.

**12.** Get the average word length per chapter.

**13.** Plot the average word length per chapter.

**14.** What is the longest word used in Harry Potter?

**15.** What is the most frequent word used in Harry Potter?

**16.** Get the number of words per chapter.

**17.** Plot the number of words per chapter.

**18.** What's the longest chapter in Harry Potter?

**19.** Run the below to create an object named `sw`.

```{r, eval = FALSE}
sw <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/stopwords.csv')
```

**20.** Remove the stop words form your one-row-per-word dataframe.

**21.** What is the most frequently used (non-stop) word in Harry Potter?

**22.** Create an object called `sentiments` by running the following:

```{r, eval = FALSE}
sentiments <- get_sentiments('afinn')
```

**23.** Use `left_join` to bring a sentiment classification to each word.

**24.** Is Harry Potter more negative or positive?

**25.** Calculate the average sentimentality per chapter.

**26.** Plot the average sentimentality per chapter.

**27.** Create a variable called `cumulative_sentiment`. Use `cumsum` to get the cumulative sum sentimentality.

**28.** Plot cumulative sentiment.

**29.** Color your plot by chapter.


