# (PART) Deep `R` {-}

# Saving data & plots

#### Learning goals {-}
- How to save dataframes as `.csv`'s
- How to save `R` data objects as `.rds`'s
- How to save plots as `.pdf`'s and `.png`'s

&nbsp;

To practice saving data (a.k.a. **exporting** data), let's use the built-in dataset from package `babynames`.

```{r,echo=TRUE,eval=FALSE}
install.packages("babynames")
library(babynames)
data(babynames)
```

```{r,echo=FALSE}
library(babynames)
data(babynames)
```

## `write.csv()` {-}

To write a dataframe to a `.csv`, use the function `write.csv()`. 

```{r,echo=TRUE,eval=FALSE}
write.csv(babynames,file="my_babynames.csv",quote=FALSE,row.names=FALSE)
```

This dataset will be saved as the file, `my_babynames.csv` within your parent directory. You can set other destinations using the same rules for child paths, parent paths, and absolute paths from the module on **Importing Data**.  

The code above sets a couple parameters that preserves the neat formatting of your dataframe. We recommend using these inputs each time you use `write.csv()` to save a dataframe.

## `saveRDS()` {-}

The `saveRDS()` function saves `R` objects in the `.rds` format, which makes them easier and faster to read into `R` later on. 

```{r,echo=TRUE,eval=FALSE}
saveRDS(babynames,file="my_babynames.rds")
```


## `pdf()` and `png()`  {-}

Let's say you want to plot the prevalence of the name "Barack" in recent U.S history:

```{r,echo=TRUE,collapse=TRUE}
head(babynames)
baracks <- babynames[babynames$name=="Barack",]
plot(prop~year,data=baracks,
     type="b",pch=16,
     xlim=c(2000,2018),
     xlab="Year",ylab="Proportion of names",main="Barack")
```

To save this as a figure as a `pdf`, preface these lines of code with a `pdf()` command and add `dev.off()` below those lines. This opens up a `pdf` 'device' (i.e., an engine for creating a PDF file), prints the plot to that new file, then closes the device and saves the file. 

Saving a pdf:

```{r,echo=TRUE,eval=FALSE}
pdf("barack-years.pdf",width-6,height=4)
plot(prop~year,data=baracks,
     type="b",pch=16,
     xlim=c(2000,2018),
     xlab="Year",ylab="Proportion of names",main="Barack")
dev.off()
```

Make sure you add '.pdf' to the end of your filename, and try not to run the `dev.off()` command when you have not first run the `pdf` command. If you do, `R` will think you are turning off the plotting feature in `RStudio` and you will have to restart your `R` session in order to see any more plots.  


