# (PART) Advanced skills {-}

# Mapping

#### Group exercise  {-}

1. Create a new rmarkdown document.

2. Create a code chunk. In this chunk, read in some data on "conflicts". This data comes from https://ucdp.uu.se/encyclopedia. Take a minute or two to look at the website. To read in the data, run the below code.

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
library(dplyr)
download.file('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/conflicts.RData',
              destfile = 'conflicts.RData')
load('conflicts.RData')
```

3. Have a look in the data. Which fields are likely to be geographic?

4. Make a simple x-y plot using geographic fields.

```{r,echo=FALSE,collapse=TRUE, eval = FALSE}
ggplot(data = conflicts,
       aes(x = longitude,
           y = latitude)) +
  geom_point()
```

5. Create an object named `conflicts_afg`. This should be a plot of conflicts in Afghanistan.

6. Color the points by year.

7. Instead of year, color the points by `deaths_civilians`.

8. Color the points by date, but make point size reflect `deaths_civilians`.

9. Create a leaflet map of conflicts in a country of your choice.

10. Use `addTiles`.

11. Use `addProviderTiles` to make your map a satellite map.

12. Explore other tiles

13. Add pop-ups to your maps by using the `popup` argument within `addMarkers`.

14. Add `clusterOptions = markerClusterOptions()` to make your points clustered.

15. Replace your markers with "circle markers".

16. Create a shiny app wherein the user selects a country and time frame, and the app shows both (a) an interactive map and (b) a plot of the number of conflicts by year for that country. 


