# Geographic computing & GIS




Let's start by getting some data. Run the below code:

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}

# This code is just for getting sewanee data into the repo in an easily readable format...
library(dplyr)
library(rgdal)
library(raster)
library(sp)
destination_directory <- '/tmp'
destination_file <- file.path(destination_directory, 'sewanee.zip')
download.file('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/sewanee.zip',
              destfile = destination_file)
unzip(destination_file, exdir = destination_directory)
boundary <- readOGR(destination_directory, 'Boundary2016')
structures <- readOGR(destination_directory, 'Domain_Structures')
roads <- readOGR(destination_directory, 'Roads')
elevation <- raster(file.path(destination_directory,
                              'DEM USGS 10m.tif'))
```



### In-class exercises  {-}


#### Raster {-}

1. What is the difference between raster data and vector data?

2. What kinds of vector data are there?

3. Let's talk about projections: https://en.wikipedia.org/wiki/List_of_map_projections

4. Let's fetch some raster data. 

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
library(raster)
library(sp)
library(rasterVis)
usa <- getData('alt', country='USA', mask=TRUE)
```

5. What kind of data is this? What is the structure?

6. Let's break it down into just the continental United States.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
cont <- usa[[1]]
plot(cont)
```

7. Plot it!


```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(cont)
```

8. What do the values mean (the legend)?

9. Make a plot of Alaska.

10. Let's retrieve some data for boundaries of States.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
states <- getData(name = 'GADM', level = 1, country = 'USA')
```

11. Plot the states.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(states)
```

12. Take a peak at the states data.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
head(states)
```

13. Make an object just for Tennessee.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
tn <- states[states$NAME_1 == 'Tennessee',]
```

14. Plot Tennessee.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(tn)
```


15. Use the `crop` and `mask` functions to get just Tennessee elevation.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
tn_elev <- crop(cont, tn)
tn_elev <- mask(tn_elev, tn)
plot(tn_elev)
```

16. Cool, yeah? Now do the same for your favorite state.

17. Make a plot of Tennessee. And then add `elevation` (Sewanee elevation) to it.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(tn)
plot(elevation, add = T)
```

Oh no! That didn't work. Why not? Have a look at coordinates

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
coordinates(tn)
coordinates(elevation)
```

18. It seems like things are on different coordinate systems. Let's convert elevation to the coordinate system of Tennessee.

```{r}
elevation_ll <- projectRaster(elevation, crs = proj4string(tn))
```

19. Great! Now we can try plotting Sewanee elevation on Tennessee again

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(tn)
plot(elevation_ll, add = T)
```


20. Make a plot of Tennessee and add Sewanee's domain (bounadry) to it.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(tn)
plot(boundary, add = T)
```

Uh oh. Same problem as before. We need to "reproject" boundary to latitude longitude.

21. Do it. Use the `spTransform` function (not `projectRaster` since `boundary` is not a raster).

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
boundary_ll <- spTransform(boundary, proj4string(tn))
plot(tn)
plot(boundary_ll, add = T)
```


22. Plot the Sewanee elevation in latitude and longitude and then add Sewanee boundary

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(elevation_ll)
plot(boundary_ll, add = T)
```

23. Use crop and mask to get just the elevation for the domain. Your plot should look like this.

```{r,echo=FALSE,collapse=TRUE, eval = TRUE}
domain_elev <- crop(elevation_ll, boundary_ll)
domain_elev <- mask(domain_elev, boundary_ll)
plot(domain_elev)
```


24. Add roads to the plot

```{r,echo=FALSE,collapse=TRUE, eval = TRUE}
plot(roads, add = T)
```

Oh no! Projection problems.

25. Reproject roads and add them.

```{r,echo=FALSE,collapse=TRUE, eval = TRUE}
roads_ll <- spTransform(roads, proj4string(domain_elev))
#plot(roads_ll, add = T)
```

26. Trim down the roads so that we only include those which are in the domain boundary area using the `over` function.

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
o <- over(roads_ll, polygons(boundary_ll))
roads_ll_trim <- roads_ll[!is.na(o),]
```

27. Make the below plot

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
plot(domain_elev)
plot(roads_ll_trim, add = TRUE)
```

28. Let's make a rasterVis contour raster plot

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
library(rasterVis)
rasterVis::contourplot(domain_elev)
```

29. Let's make a level plot

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
rasterVis::levelplot(domain_elev)
```

30. Let's make a level plot

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
rasterVis::levelplot(domain_elev)
```

31. How about a ggplot2-style plot

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
library(ggplot2)
domain_elev_df <- as.data.frame(domain_elev, xy = TRUE) %>%
  filter(!is.na(DEM_USGS_10m))


ggplot(data = domain_elev_df,
       aes(x = x,
                  y = y,
                  fill = DEM_USGS_10m)) +
  geom_raster() +
  scale_fill_gradient2(low = 'white', high = 'red')
```

32. Make a ggplot-style raster plot of USA elevation


33. Let's "bin" values in `elevation` to say "high" or "low".

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
cols <- c('black', 'yellow')
# add breaks to the colormap (6 breaks = 5 segments)
brk <- c(0, 300, 1000)

plot(elevation, col=cols, breaks=brk, main="High vs low")
```


34. Do the same as above, but make 3 colors.


35. Let's make a leaflet raster!

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
library(leaflet)
pal <- colorNumeric(c("#0C2C84", "#41B6C4", "#FFFFCC"), values(elevation_ll),
  na.color = "transparent")
leaflet() %>% 
  addTiles() %>%
  addRasterImage(elevation_ll, colors = pal, opacity = 0.8) %>%
  addLegend(pal = pal, values = values(elevation_ll),
    title = "Elevation")
```

36. Add `structures` to the above. Hint: you'll need to use `addPolygons` and you'll need to reproject structures as `structures_ll`...

37. Add popups to your structures.

38. Make your structures a different color and remove the border (hint, you'll need to use the `stroke` argument)

39. Get the elevation of each structure by running:

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
structure_elevation <- 
  unlist(lapply(extract(elevation_ll, structures_ll),
         function(x){
           mean(x,na.rm = TRUE)
         }))
```

40. Use the `structure_elevation` object to add a new column to the `structures_ll` object.

41. Make a histogram of the elevation of Sewanee buildings.

42. How low is the lowest building on the domain? How low is it?

43. How high is the highest building on the domain? Which building is it?


#### Shapefiles / polygons {-}


```{r,echo=FALSE,collapse=TRUE, eval = FALSE}

# This code is just for getting sewanee data into the repo in an easily readable format...
library(dplyr)
library(rgdal)
library(raster)
library(sp)
destination_directory <- '/tmp'
destination_file <- file.path(destination_directory, 'world_shp.zip')
download.file('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/world_shp.zip',
              destfile = destination_file)
unzip(destination_file, exdir = destination_directory)
world_shp <- readOGR(file.path(destination_directory, 'world_small/'),
                     'TM_WORLD_BORDERS_SIMPL-0.3')

save(world_shp, file = 'data/world_shp.RData')
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
file_source <- 'https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/world_shp.RData'
library(dplyr)
library(rgdal)
library(raster)
library(sp)
library(readr)
destination_directory <- '/tmp'
destination_file <- file.path(destination_directory, 'world_shp.RData')
if(!'data/world_shp.RData' %in% destination_directory){
  download.file(file_source,
                destfile = destination_file)
}
load(destination_file)

# Read in indicator data
df <- read_csv('https://raw.githubusercontent.com/databrew/intro-to-data-science/main/data/hefpi.csv')

shp <- world_shp
```


Subset data by indicator & join with shape file data

```{r}
pd <- df %>% filter(indicator_name =='Inpatient care use, adults')
shp@data <- left_join(shp@data, pd)
```

Make a basic (ugly) map 

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
library(leaflet)
library(RColorBrewer)

# map text
map_palette <- colorNumeric(palette = brewer.pal(9, "Greens"), domain=shp@data$value, na.color="#CECECE")

    leaflet(shp) %>% 
        addProviderTiles(provider = providers$Esri.WorldShadedRelief) %>%
        addPolygons( 
          fillColor = ~map_palette(value),
          fillOpacity = 0.9)
        

```

Set min and max zoom

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}

    leaflet(shp, options = leafletOptions(minZoom = 1,maxZoom = 10)) %>% 
        addProviderTiles(provider = providers$Esri.WorldShadedRelief) %>%
        addPolygons( 
          fillColor = ~map_palette(value),
          fillOpacity = 0.9)
        

```


Add colors based on value

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}

    leaflet(shp, options = leafletOptions(minZoom = 1,maxZoom = 10)) %>% 
        addProviderTiles(provider = providers$Esri.WorldShadedRelief) %>%
        addPolygons( 
          color = 'black',
          weight=1,
          fillColor = ~map_palette(value), 
          stroke=TRUE,
          fillOpacity = 0.9, 
          label = ~round(value, 2))
        

```

Add legend

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}

    leaflet(shp, options = leafletOptions(minZoom = 1,maxZoom = 10)) %>% 
        addProviderTiles(provider = providers$Esri.WorldShadedRelief) %>%
        addPolygons( 
          color = 'black',
          weight=1,
          fillColor = ~map_palette(value), 
          stroke=TRUE,
          fillOpacity = 0.9, 
          label = ~country
        ) %>% 
        addLegend( pal=map_palette, values=~value, opacity=0.9, position = "bottomleft", na.label = "NA" )

```

Add fancy text to map

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
library(htmltools)

# Create map
map_text <- paste(
  "Indicator: ",  shp@data$indicator_name,"<br>",
  "Economy: ", as.character(shp@data$country),"<br/>", 
  'Value: ', round(shp@data$value, digits = 2),  "<br/>",
  "Year: ", as.character(shp@data$year),"<br/>",sep="") %>%
  lapply(htmltools::HTML)

    leaflet(shp, options = leafletOptions(minZoom = 1, 
                                                maxZoom = 10)) %>% 
        addProviderTiles('Esri.WorldShadedRelief') %>%
        addPolygons( 
          color = 'black',
          fillColor = ~map_palette(value), 
          stroke=TRUE, 
          fillOpacity = 0.9, 
          weight=1,
          label = map_text,
          highlightOptions = highlightOptions(
            weight = 1,
            fillColor = 'white',
            fillOpacity = 1,
            color = "white",
            opacity = 1.0,
            bringToFront = TRUE,
            sendToBack = TRUE
          ),
          labelOptions = labelOptions( 
            noHide = FALSE,
            style = list("font-weight" = "normal", padding = "3px 8px"), 
            textsize = "13px", 
            direction = "auto"
          )
        ) %>% 
        addLegend( pal=map_palette, values=~value, opacity=0.9, position = "bottomleft", na.label = "NA" )

```

## Exercise {-}

1. Make a choropleth map of BMI for men, where the darker the shade of red, the higher the BMI for each country. 

2. Remove the borders from the map

3. Add a legend on the top right of the map

4. Make the NA color blue

5. Make the hover label a combination of the country and BMI value

6. Make the title of the legend "BMI" 

7. Create a function that takes an indicator name as an input and creates a map.

