# Working with text {#text}

#### Learning goals {-}

* Learn to apply the most common `R` tools for working with text.


## Basics {-}

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
library(dplyr)
library(gsheet)
survey <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1iVt9FX9J2iv3QFKBM7Gzb9dgva70XrW1lxMV4hpekeo/edit?resourcekey#gid=204634767')
names(survey) <- c('time', 'sex', 'age','sib', 'dad_mus', 'person_mus', 'joe_mus_is', 'eyesight', 'height', 'shoe_size', 'bday', 'money_or_love', 'rps_skill', 'num_pan', 'cats_dogs', 'first_name', 'last_name')

```

#### `paste0()` {-}

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# combine first name and last name 
paste0(survey$first_name,' ', survey$last_name)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
paste0(survey$first_name, ' ', survey$sex)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# Make a sentence
paste0('First name is ', survey$first_name,' and last name is  ', survey$last_name)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# combine first and last name to make new variable called full_name
survey <- survey %>% mutate(full_name = paste0(survey$first_name, '_', survey$last_name ))
```

#### `tolower()` & `toupper()` {-}

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# combine first name and last name 
tolower(survey$money_or_love)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
toupper(survey$money_or_love)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# overwrite money_or_love
survey <- survey %>% mutate(money_or_love = tolower(money_or_love))
```

#### `nchar()` {-}

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# count how many characters are in each observations in sex
nchar(survey$eyesight)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# create a boolean vector of the number of observations with 32 characters
nchar(survey$eyesight) ==42
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# get number of observations with character 42
length(which(nchar(survey$eyesight) ==42))
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
sum(nchar(survey$eyesight) == 42)
```

##### `substr()` {-}

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
# grab only the first letter fo sex
substr(survey$sex, start = 1, stop = 1)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# remove the "s" from cats and dogs
substr(survey$cats_dogs, start = 1, stop = 3)
```

#### `gsub()` {-}

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
# replace "Deeply captivating" with "captivating
gsub('Deeply captivating', 'captivating', survey$joe_mus_is)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# remove long strings
gsub('(glasses or contacts, but I can get by without them)','', survey$eyesight, fixed = TRUE)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# overwrite variable
survey <- survey %>% mutate(eyesight = gsub('(glasses or contacts, but I can get by without them)','', eyesight, fixed = TRUE))
```

#### `grepl()` {-}

```{r,echo=TRUE,collapse=TRUE, eval = FALSE}
# how many times does Brew show up in full_name
grepl('Brew', survey$full_name)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# Born in may
grepl('05', survey$bday)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# Born in 2000
grepl('2000', survey$bday)
```

```{r,echo=TRUE,collapse=TRUE, eval = TRUE}
# filter data by those born in 2000
survey %>% filter(grepl('2000', bday))
```

#### Exercises {-} 

1) Create a new variable called `first_name_sex` that combines the first name and the sex of each individual.

2) Create a new variable called `backwards_name` that combines last name and first name. 

3) Make the `backwards_name` variable capitalized.

4) Replace "both" in the `money_or_love` variable with "Money & Love".

5) Get only the first character `dad_mus` variable. 

6) How many total characters are in the column `eyesight`?

7) How many characters did Joe write for the `eyesight` question?

8) How many people in the data were born on the 4th day of the month?

9) Create a new variables called `month_born` that has only the month of from the `bday` variable. 

10) Do the same thing for `year`. 

11) Filter the data set by those born and 2001 and prefer money.


