--- 
title: "`DataBOOM`: the canon for data science"
author: "Databrew"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: databrew/intro-to-data-science
description: "Data science is just data and science. But it's better with pizza."
params:
  is_teacher: TRUE
---

```{r setup, include=FALSE, echo = FALSE}
# Basic knitr options
library(knitr)
opts_chunk$set(comment = NA, 
               echo=FALSE,
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE,
               fig.width = 9.64,
               fig.height = 5.9,
               fig.path = 'figures/')
options(scipen=999)
is_teacher <- params$is_teacher

```

```{r,eval=TRUE,echo=FALSE}
# Define function for showing teacher content
teacher_content <- function(content = ''){
  cat(paste0('<p style="color:#3f0000; background-color:#f5f5dc"> <b>Instructor tip!</b><br>', content, '</p>'))
}
```

# Welcome!

Welcome to  ``DataBOOM``, a curriculum designed to guide you from your very first line of code towards becoming a professional data scientist. 


## What this is, and what it isn't {-}

This is not a textbook or a reference manual. It is not exhaustive or comprehensive. It is a *training manual* designed to *empower researchers to do impactful data science.* As such, its tutorials and exercises aim to get you, the researcher, to start writing your own code as quickly as possible and -- equally of importance -- to *start thinking like a data scientist*, by which we mean tackling ambiguous problems with persistence, independence, and creative problem solving.  

Furthermore, this is not a fancy interactive tutorial with bells or whistles. It was purposefully designed to be simple and "analog". You will not be typing your code into this website and getting feedback from a robot, or setting up an account to track your progress, or getting pretty merit badges or points when you complete each module.  

Instead, you will be doing your work on your own machine, working with real folders and files, downloading data and moving it around, etc. -- all the things you will be doing as a data scientist in the real world.


## Who this is for {-}

This curriculum covers everything from the absolute basics of writing code in `R` to machine learning with `tensorflow`. As such, it is designed to be useful to everyone in some way. But the target audience for these tutorials is the student who *wants* to work with data but has *zero* formal training in programming, computer science, or statistics.  

This curriculum was originally developed for the **Sewanee Data Institute for Social Good** at Sewanee: The University of the South, TN.


## What you will learn {-}

- The **Core theory** unit establishes the conceptual foundations and motivations for this work: what data science is, why it matters, and ethical issues surrounding it: the good, the bad, and the ugly.

The next several units comprise a *core* curriculum for tackling data science problems:  

- The **Getting started** unit teaches you how to use `R` (in `RStudio`) to explore and plot data. Here you will add the first and most important tools to your toolbox: working with variables, vectors, dataframes, scripts, and file directories.

- The **Working with data** unit teaches you how to bring in your own data and work with it in `R`. You will learn how to format data to simplify analysis and add tools for *data wrangling* (i.e., transforming and re-formatting data to prepare it for plotting and analysis).

- The **Exploring & analyzing data** unit teaches you how to conduct basic statistics, from exploratory data analyses (e.g., producing and comparing distributions) to significance testing.  

- The **Accessing & managing data** unit teaches you how to organize data for large projects, how to manage atypical data formats (e.g., Excel files), and how to access data from online databases.

- The **`R` toolbag** unit equips you with the tools, tricks, and mindset for tackling the most common tasks in data science. This is where you really begin to cut your teeth on real-world data puzzles: figuring out how to use the `R` tools in your toolbag to tackle an ambiguous problem and deliver an excellent data product. 

The next several units provide a suite of skills essential to any data science professional:

- The **Interactive dashboards** unit teaches you how to make dashboards and websites for projects using `shiny` in `RStudio`.  

- The **Databases** unit teaches you how to access, create, and work with online databases using `SQL` and its alternatives.  

- The **Documenting your work** unit teaches you to use `R Markdown` to produce beautiful, reproducible data reports.  

- The **Version control & teamwork** unit teaches you how to use `Git` and `GitHub` to collaborate on shared projects and work on data science teams.  

- The **Writing about data** unit teaches you to write successful grant applications and publishable research articles.  


The final unit, **Advanced skills**, introduces you to a variety of advanced data science techniques, from interactive maps to iterative simulations to machine learning, that can help you begin to specialize your skillset.  


## Making the most of this course   {-}

(Make sure you involve pizza.)  


## Who we are {-}

Joe Brew  

Ben Brew  

Eric Keen  

Isabelle Puckette  





```{r, results = 'asis', eval = is_teacher}
teacher_content('Here is some teacher content.')
```



```{r,include=TRUE,eval=FALSE}
library(dplyr)
library(Hmisc)
library(knitr)
files <- dir()
files <- files[grepl('.Rmd', files)]
files <- files[files != 'index.Rmd']
files <- gsub('.Rmd', '', files)
files <- Hmisc::capitalize(files)
files <- gsub('_', ' ', files)
files_split <- strsplit(files, split = '-')
chapter_numbers <- unlist(lapply(files_split, function(x){x[1]}))
chapter_names <- unlist(lapply(files_split, function(x){x[2]}))
toc <-
  dplyr::tibble(
    Number = chapter_numbers,
    Name = chapter_names
  )
toc <- toc[which(toc$Number != "Intro"),]
knitr::kable(toc)
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
