--- 
title: "Data Science Boot Camp"
author: "Databrew"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: databrew/intro-to-data-science
description: "Data science is just data and science."
---

```{r setup, include=FALSE, echo = FALSE}
# Basic knitr options
library(knitr)
opts_chunk$set(comment = NA, 
               echo=FALSE,
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE,
               fig.width = 9.64,
               fig.height = 5.9,
               fig.path = 'figures/')
options(scipen=999)
```


# Welcome

Welcome to the Data Science Boot Camp.

```{r}
library(dplyr)
library(Hmisc)
library(knitr)
files <- dir()
files <- files[grepl('.Rmd', files)]
files <- files[files != 'index.Rmd']
files <- gsub('.Rmd', '', files)
files <- Hmisc::capitalize(files)
files <- gsub('_', ' ', files)
files_split <- strsplit(files, split = '-')
chapter_numbers <- unlist(lapply(files_split, function(x){x[1]}))
chapter_names <- unlist(lapply(files_split, function(x){x[2]}))
toc <-
  dplyr::tibble(
    Number = chapter_numbers,
    Name = chapter_names
  )
knitr::kable(toc)
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!--chapter:end:index.Rmd-->

# Introduction {#intro}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

<!--chapter:end:01-intro.Rmd-->

# Literature

Here is a review of existing methods.

<!--chapter:end:02-literature.Rmd-->

# Methods

We describe our methods in this chapter.

<!--chapter:end:03-method.Rmd-->

# Applications

Some _significant_ applications are demonstrated in this chapter.

## Example one

## Example two

<!--chapter:end:04-application.Rmd-->

# Final Words

We have finished a nice book.

<!--chapter:end:05-summary.Rmd-->

`r if (knitr:::is_html_output()) '
# References {-}
'`

<!--chapter:end:06-references.Rmd-->

